### 2016-07-12 30June16_plate_3_explant_sup.Rmd
First commit

---
title: '30June16 explant sups plate 3: MIG, RANTES, IL1ra'
author: "Claire Levy"
date: "July 6, 2016"
output: github_document

---
This is an analysis of data from running our MSD plate 3 on some of HSV2 infected explant sup samples for Herpes P01. See explant_and_HVE_cell_sup_MSD_README for more on the different plates/analyte panels. This plate has the following analytes measured in each well

Plate 3: 96 well 4 spot prototype Human Triplex

	Il-1ra
	MIG (CXCL9)
	RANTES (CCL5)

The MSD machine gives the data in an excel or text format. I have .csv file with the results.

```{r setup, echo=FALSE, warning=FALSE, message= FALSE}
library(dplyr)
library(ggplot2)
library(pander)
library(reshape2)
library(stringr)
library(swarm)

#read in the .csv raw results file
#../ to go up one level and then back down into a different folder. I exported this file from the DWB experiment called "Hladik plate 3 30 June 16 corrected layout". I imported that file from the folder location where FC saved the DWB experiment with the corrected layout. That file is here:

#J:\MacLabUsers\HLADIK SHARED\Projects\Herpes study\MSD\CVL Test Samples\Raw Data from MSD\Test Run

#I appended the name with "corrected layout" because I imported the wrong file at first and you can't delete them from the software.

dat<- read.csv("../raw_data/30June16_plate3_CVL_and_explant_sup_CLexport.csv", skip = 1)

#### NOTE ####
# This file also contains data from TEST CVL samples from the hvtn clinic that we were running to test the analytes before doing the actual trial samples from Kenya.

#make a column identifying what group the sample belongs to, standards or explant sup or CVL sample so I can filter out what I want.

dat<- dat %>%
  mutate(Type =
 ifelse(str_detect(Sample,"^3")==TRUE, "explant_sup",
        ifelse(str_detect(Sample, "R|V")==TRUE,
                          "CVL","Standard")))

## Also, we included the diluted stock calibrator and the blended working stock on the plate when we shouldn't have. So we excluded those samples in the Discovery workbench software, which re-sets the curve fit based on the remaining samples. When you export the data table, it still includes those EXCLUDED samples. I am removing them here.

dat <- dat %>%
  filter(Sample != "Diluted Stock Calibrator" & Sample != "Blended Working Stock")
```

```{r standards, echo=FALSE, warning=FALSE, message= FALSE}

#filtering out the standards
standards <- dat %>%
  filter(Type == "Standard")

#which standards are in the detection range
standardsInRange <-standards%>%
  filter(Detection.Range == "In Detection Range")

#which were not in the detection range?

standardsOutOfRange<-standards%>%
  filter(Detection.Range != "In Detection Range")%>%
  arrange(Type, Assay, Sample)%>%
  select(Type,Sample, Assay, Detection.Range, X..Recovery)

```

Of the `r length(standards)` standards that we measured, `r length(standardsInRange)` are in the detection range. The following standards were out of the detection range:
`r pander(standardsOutOfRange)`

Here is a plot of the % recovery (extrapolated concentration/ known concentration x 100) for the standards that were in the detection range, with lines at the "good recovery" limits of 80 and 120% recovery.

```{r percent recovery, echo=FALSE, warning=FALSE, message= FALSE}

stdFit<- standardsInRange %>% 
  arrange(Sample,Assay)%>%
  select(Type, Sample, Assay, X..Recovery,Fit.Statistic..RSquared, Concentration, Calc..Concentration)

ggplot(stdFit, aes(x = Sample, y = X..Recovery))+
  geom_point(aes(color = Sample))+
  facet_wrap(~Assay)+
  theme(axis.text.x = element_text(angle=25))+
  scale_y_continuous(limits = c(75, 200))+
  geom_abline(slope = 0, intercept = c(80, 120))+
  labs(y = "Percent Recovery")

```

The R-squared values for goodness-of-fit were good for all assays:

```{r r-squared, echo=FALSE, warning=FALSE, message= FALSE}

rSquared<-stdFit%>%
  group_by(Assay)%>%
  summarize(R_Squared = mean(Fit.Statistic..RSquared))

pander(rSquared)
```

```{r explant data cleanup, echo=FALSE, warning=FALSE, message= FALSE}

#filtering for the samples I want (just want explant sups, not CVLS or standards)

explantDat<-dat%>%
  filter(Type == "explant_sup")

#sanity check: I put 56 samples on the plate and there are 3 analytes so we should have 56 x 3 = 168 samples.

#We didn't have enough sample to do two replicates for 324-T2-M so we didn't add anything to well F10. We did read the well though, so there is data for it, but I want to exclude it because there wasn't sample there

explantDat <- explantDat %>%
  filter(Well != "F10") #removes 3 data points, 1 for each analyte so 165 left

#which explant samples are in the detection range
explantDatInRange <-explantDat%>%
  filter(Detection.Range == "In Detection Range")

#which were not in the detection range?

explantDatOutOfRange<-explantDat%>%
  filter(Detection.Range != "In Detection Range")%>%
  arrange(Type, Assay, Sample)%>%
  select(Type,Sample, Assay, Detection.Range)

```

```{r adding ids and avg reps, echo=FALSE, warning=FALSE, message= FALSE}

# I want to be able to group the data by timepoint and donor and virus so I need to separate the sample name and make new columns identifying them that way

donorTimeVirus<-colsplit(explantDat$Sample,"-", c("Donor","Timepoint","Virus") )

#bind the split col back to explantDat

explantDat<-cbind(donorTimeVirus, explantDat)

#I should only have 1 replicate for 324-T2-M. All other samples have 2 reps.

#There is one sample where the concentration was below the fit curve range and the concentration is given as "NaN". In order for that point to appear on the plot, I am going to give those samples a value of "1"

NaNSamples<- explantDat %>%
  filter(Calc..Concentration == "NaN")%>%
  select(Sample, Detection.Range, Calc..Concentration)

explantDat <- explantDat %>%
  mutate(Calc..Concentration = ifelse(Calc..Concentration == "NaN", "1", Calc..Concentration))

explantDat$Donor <- as.factor(explantDat$Donor)

explantDat$Calc..Concentration <-as.numeric(explantDat$Calc..Concentration)

explantDat$Type <- as.factor(explantDat$Type)

explantDat$Virus <- as.factor(explantDat$Virus)

explantDat$Virus<-factor(explantDat$Virus, levels = c("M", "V1","V2","M 1:50", "V2 1:50"))

explantDat$Timepoint<-as.factor(explantDat$Timepoint)

#Averaging the replicates. 
#Sanity check: 
#For 27/28 samples I have 2 replicates = 54 
# For 1/28 samples I have 1 replicate = 1
#If I average those samples, I get 54/2 + 1 = 28 avg values
# I have data for 3 analytes so I should have 28 * 3 average values.

AvgConcExplantDat <- explantDat %>%
  group_by(Sample, Assay, Donor, Timepoint,Virus, Detection.Range)%>%
  summarize(avgConc = mean(Calc..Concentration))
```

MIG data

```{r MIG, echo=FALSE, warning=FALSE, message= FALSE}

#I want to make separate plots for each analyte, so I'll split it into different dfs here:

#MIG

MIGAvgConcExplantDat<- AvgConcExplantDat %>%
  filter(Assay == "MIG")

ggplot(data = MIGAvgConcExplantDat, aes(x = Virus, y = avgConc))+

  geom_point(aes(color = Donor, shape = Detection.Range ),size = 2 )+

  geom_line(aes(color = Donor, group = Donor))+

  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+

  scale_y_log10()+

  facet_wrap(~Timepoint)+

  theme(axis.text.x = element_text(angle = 25))
```

RANTES data

```{r RANTES, echo=FALSE, warning=FALSE, message= FALSE}
#RANTES

RANTESAvgConcExplantDat<- AvgConcExplantDat %>%
  filter(Assay == "RANTES")

ggplot(data = RANTESAvgConcExplantDat, aes(x = Virus, y = avgConc))+

  geom_point(aes(color = Donor, shape = Detection.Range ),size = 2 )+

  geom_line(aes(color = Donor, group = Donor))+

  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+

  scale_y_log10()+

  facet_wrap(~Timepoint)+

  theme(axis.text.x = element_text(angle = 25))
```

IL1-RA data

```{r IL1-ra, echo=FALSE, warning=FALSE, message= FALSE}
## IL1-ra

IL1raAvgConcExplantDat<- AvgConcExplantDat %>%
  filter(Assay == "IL1-RA")

ggplot(data = IL1raAvgConcExplantDat, aes(x = Virus, y = avgConc))+

  geom_point(aes(color = Donor, shape = Detection.Range ),size = 2 )+

  geom_line(aes(color = Donor, group = Donor))+

  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+

  scale_y_log10()+

  facet_wrap(~Timepoint)+

  theme(axis.text.x = element_text(angle = 25))

```

### 2016-07-12 30June16_plate_3_explant_sup.Rmd
Added plot titles, fixed inline code

@@
 output: github_document
-
 ---
@@
 
-The MSD machine gives the data in an excel or text format. I have .csv file with the results.
-
 ```{r setup, echo=FALSE, warning=FALSE, message= FALSE}
@@
 library(swarm)
+
+#The MSD machine gives the data in an excel or text format. I have .csv file with the results.
 
@@
   arrange(Type, Assay, Sample)%>%
-  select(Type,Sample, Assay, Detection.Range, X..Recovery)
+  select(Sample, Assay, Detection.Range)
 
@@
 
-Of the `r length(standards)` standards that we measured, `r length(standardsInRange)` are in the detection range. The following standards were out of the detection range:
+Of the `r nrow(standards)` standards that we measured, `r nrow(standardsInRange)` are in the detection range. The following standards were out of the detection range:
 `r pander(standardsOutOfRange)`
@@
   geom_abline(slope = 0, intercept = c(80, 120))+
-  labs(y = "Percent Recovery")
+  labs(y = "Percent Recovery")+
+  theme(axis.text.x = element_text(size = rel(0.8)))
 
@@
 
-MIG data
-
 ```{r MIG, echo=FALSE, warning=FALSE, message= FALSE}
@@
 
-MIGAvgConcExplantDat<- AvgConcExplantDat %>%
+MIGexplantDat<- explantDat %>%
   filter(Assay == "MIG")
 
-ggplot(data = MIGAvgConcExplantDat, aes(x = Virus, y = avgConc))+
+ggplot(data = MIGexplantDat, aes(x = Virus, y = Calc..Concentration))+
 
-  geom_point(aes(color = Donor, shape = Detection.Range ),size = 2 )+
+  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5 )+
 
-  geom_line(aes(color = Donor, group = Donor))+
+  stat_summary(aes(group = Donor, color = Donor),fun.y = mean, geom = "point", size = 2)+
+
+  stat_summary(aes(group = Donor, color = Donor), fun.y =mean, geom = "line")+
 
@@
 
+  labs( y = "Concentration")+
+
   facet_wrap(~Timepoint)+
 
-  theme(axis.text.x = element_text(angle = 25))
+  theme(axis.text.x = element_text(angle = 25))+
+  ggtitle("MIG data")
+
 ```
-
-RANTES data
 
@@
 
-RANTESAvgConcExplantDat<- AvgConcExplantDat %>%
+RANTESexplantDat<- explantDat %>%
   filter(Assay == "RANTES")
 
-ggplot(data = RANTESAvgConcExplantDat, aes(x = Virus, y = avgConc))+
+ggplot(data = RANTESexplantDat, aes(x = Virus, y = Calc..Concentration))+
 
-  geom_point(aes(color = Donor, shape = Detection.Range ),size = 2 )+
+  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5 )+
 
-  geom_line(aes(color = Donor, group = Donor))+
+  stat_summary(aes(group = Donor, color = Donor),fun.y = mean, geom = "point", size = 2)+
+
+  stat_summary(aes(group = Donor, color = Donor), fun.y =mean, geom = "line")+
 
@@
 
+  labs( y = "Concentration")+
+
   facet_wrap(~Timepoint)+
 
-  theme(axis.text.x = element_text(angle = 25))
+  theme(axis.text.x = element_text(angle = 25))+
+  ggtitle("RANTES data")
+
 ```
-
-IL1-RA data
 
@@
 
-IL1raAvgConcExplantDat<- AvgConcExplantDat %>%
+IL1RAexplantDat<- explantDat %>%
   filter(Assay == "IL1-RA")
 
-ggplot(data = IL1raAvgConcExplantDat, aes(x = Virus, y = avgConc))+
+ggplot(data = IL1RAexplantDat, aes(x = Virus, y = Calc..Concentration))+
 
-  geom_point(aes(color = Donor, shape = Detection.Range ),size = 2 )+
+  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5 )+
 
-  geom_line(aes(color = Donor, group = Donor))+
+  stat_summary(aes(group = Donor, color = Donor),fun.y = mean, geom = "point", size = 2)+
+
+  stat_summary(aes(group = Donor, color = Donor), fun.y =mean, geom = "line")+
 
@@
 
+  labs( y = "Concentration")+
+
   facet_wrap(~Timepoint)+
 
-  theme(axis.text.x = element_text(angle = 25))
+  theme(axis.text.x = element_text(angle = 25))+
+  ggtitle("IL1-RA data")
 

### 2016-07-12 30June16_plate_3_explant_sup.Rmd
Removed "below curve fit" data from plots

@@
 ---
-This is an analysis of data from running our MSD plate 3 on some of HSV2 infected explant sup samples for Herpes P01.
-See explant_and_HVE_cell_sup_MSD_README for more on the different plates/analyte panels.
+This is an analysis of data from running our MSD plate 3 on some of the HSV2 infected explant sup samples for Herpes P01.
+See "explant_and_HVE_cell_sup_MSD_README.txt"" for more on the different plates/analyte panels.
 This plate has the following analytes measured in each well
@@
 #There is one sample where the concentration was below the fit curve range and the concentration is given as "NaN".
-In order for that point to appear on the plot, I am going to give those samples a value of "1"
-
-NaNSamples<- explantDat %>%
-  filter(Calc..Concentration == "NaN")%>%
-  select(Sample, Detection.Range, Calc..Concentration)
+This is because values can't be extrapolated below the curve.
+I am removing these data points.
 
 explantDat <- explantDat %>%
-  mutate(Calc..Concentration = ifelse(Calc..Concentration == "NaN", "1", Calc..Concentration))
+  filter(Detection.Range != "Below Fit Curve Range")
 
@@
 
-AvgConcExplantDat <- explantDat %>%
-  group_by(Sample, Assay, Donor, Timepoint,Virus, Detection.Range)%>%
-  summarize(avgConc = mean(Calc..Concentration))
+# AvgConcExplantDat <- explantDat %>%
+#   group_by(Sample, Assay, Donor, Timepoint,Virus, Detection.Range)%>%
+#   summarize(avgConc = mean(Calc..Concentration))
 ```
@@
   facet_wrap(~Timepoint)+
+
+  theme(axis.text.x = element_text(angle = 25))+
+  ggtitle("MIG data")
+
+```
+
+Another way of looking at the data
+
+```{r MIG alternate, echo=FALSE, warning=FALSE, message= FALSE}
+
+ggplot(data = MIGexplantDat, aes(x = Timepoint, y = Calc..Concentration))+
+
+  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5 )+
+
+  stat_summary(aes(group = Donor, color = Donor),fun.y = mean, geom = "point", size = 2)+
+
+  #stat_summary(aes(group = Donor, color = Donor), fun.y =mean, geom = "line")+
+
+  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "In Detection Range" = 16))+
+
+  scale_y_log10()+
+
+  labs( y = "Concentration")+
+
+  facet_wrap(~Virus)+
 
@@
 
+```{r RANTES alternate, echo=FALSE, warning=FALSE, message= FALSE}
+
+ggplot(data = RANTESexplantDat, aes(x = Timepoint, y = Calc..Concentration))+
+
+  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5 )+
+
+  stat_summary(aes(group = Donor, color = Donor),fun.y = mean, geom = "point", size = 2)+
+
+  #stat_summary(aes(group = Donor, color = Donor), fun.y =mean, geom = "line")+
+
+  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+
+
+  scale_y_log10()+
+
+  labs( y = "Concentration")+
+
+  facet_wrap(~Virus)+
+  ggtitle("RANTES data")
+
+```
+
 ```{r IL1-ra, echo=FALSE, warning=FALSE, message= FALSE}
-## IL1-ra
 
@@
 ```
+
+```{r IL1RA alternate, echo=FALSE, warning=FALSE, message= FALSE}
+
+ggplot(data = IL1RAexplantDat, aes(x = Timepoint, y = Calc..Concentration))+
+
+  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5 )+
+
+  stat_summary(aes(group = Donor, color = Donor),fun.y = mean, geom = "point", size = 2)+
+
+  #stat_summary(aes(group = Donor, color = Donor), fun.y =mean, geom = "line")+
+
+  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+
+
+  scale_y_log10()+
+
+  labs( y = "Concentration")+
+
+  facet_wrap(~Virus)+
+  ggtitle("IL1RA data")
+```

### 2016-07-15 30June16_plate_3_explant_sup.Rmd
formatted the individual plots differently, added a plot with all analytes

@@
 dat<- read.csv("../raw_data/30June16_plate3_CVL_and_explant_sup_CLexport.csv", skip = 1)
+
+### Formatting note: In the data table, there is both " T3" and "T3" so I am fixing it here.
+
+dat$Sample<- str_replace(dat$Sample, " T3", "T3")
 
@@
 
-Here is a plot of the % recovery (extrapolated concentration/ known concentration x 100) for the standards that were in the detection range, with lines at the "good recovery" limits of 80 and 120% recovery.
+Here is a plot of the % recovery (extrapolated concentration/ known concentration x 100) for the standards that were either within or above the detection range, with lines at the "good recovery" limits of 80 and 120% recovery.
 
@@
 
-stdFit<- standardsInRange %>% 
+stdFit<- standards %>% 
   arrange(Sample,Assay)%>%
-  select(Type, Sample, Assay, X..Recovery,Fit.Statistic..RSquared, Concentration, Calc..Concentration)
+  filter(Detection.Range != "Below Fit Curve Range")%>%
+  select(Type, Sample, Assay,Detection.Range, X..Recovery,Fit.Statistic..RSquared, Concentration, Calc..Concentration)
 
 ggplot(stdFit, aes(x = Sample, y = X..Recovery))+
-  geom_point(aes(color = Sample))+
+  geom_point(aes(color = Sample, shape = Detection.Range ),size = 3)+
+  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "In Detection Range" = 16))+
   facet_wrap(~Assay)+
@@
 pander(rSquared)
+```
+
+```{r standard curves, echo=FALSE, warning=FALSE, message= FALSE}
+
+# standardCurve<- function (Calc.Top, Calc.Bottom, Calc.HillSlope, Calc.Midpoint, Concentration){
+# 
+# Calc.Top + (Calc.Bottom - Calc.Top/(1+(Concentration/Calc.Midpoint)^Calc.HillSlope))
+# }
+# 
+# 
+# MIGcurveData <- standards%>%
+#   filter(Assay == "MIG") %>%
+#   mutate(UpperLimit = Algorithm.Parameter..Calc..Top-( Detection.Limits..Top.of.Range.Parameter * Std..Deviation ))%>%
+#   mutate(LowerLimit = min(Concentration)+(2.5 * Std..Deviation ))
+# 
+# 
+# 
+# MIGcalcCurve <- standardCurve(MIGcurveData$Algorithm.Parameter..Calc..Top,MIGcurveData$Algorithm.Parameter..Calc..Bottom, MIGcurveData$Algorithm.Parameter..Calc..HillSlope, MIGcurveData$Algorithm.Parameter..Calc..MidPoint, MIGcurveData$Concentration)
+# 
+# 
+# ggplot(MIGcurveData, aes(x = Concentration, y = Signal))+
+#   geom_point()+
+#   scale_y_log10()+
+#   scale_x_log10()+
+#   geom_line(aes(y = MIGcalcCurve))+
+#   geom_hline(yintercept = MIGcurveData$UpperLimit)
+
 ```
@@
 ```
+## Caveats
+
+Some data points were above the range of the curve fit.
+The MSD software can extrapolate values for these if the curve is linear at the top.
+Values cannot be extrapolated when the data was below the fit curve range, so those are not shown in the following plots.
+
+```{r everything, echo=FALSE, warning=FALSE, message= FALSE}
+
+ggplot(explantDat, aes(x = Virus, y = Calc..Concentration))+
+
+  geom_point(aes(color = Assay, shape = Detection.Range ),size = 1.5)+
+
+  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+
+
+  scale_y_log10()+
+
+  labs( y = "Concentration")+
+
+  facet_wrap(Timepoint~Donor, scales = "free_x")+
+
+  theme(axis.text.x = element_text(angle = 25, size = 9))+
+
+  ggtitle("Data from all analytes")
+
+```
 
@@
 
-  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5 )+
+  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5, position = position_swarm(0.5) )+
 
@@
 
-  facet_wrap(~Timepoint)+
-
-  theme(axis.text.x = element_text(angle = 25))+
-  ggtitle("MIG data")
-
-```
-
-Another way of looking at the data
-
-```{r MIG alternate, echo=FALSE, warning=FALSE, message= FALSE}
-
-ggplot(data = MIGexplantDat, aes(x = Timepoint, y = Calc..Concentration))+
-
-  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5 )+
-
-  stat_summary(aes(group = Donor, color = Donor),fun.y = mean, geom = "point", size = 2)+
-
-  #stat_summary(aes(group = Donor, color = Donor), fun.y =mean, geom = "line")+
-
-  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "In Detection Range" = 16))+
-
-  scale_y_log10()+
-
-  labs( y = "Concentration")+
-
-  facet_wrap(~Virus)+
+  facet_wrap(Timepoint ~ Donor, scales = "free_x")+
 
@@
 
-  facet_wrap(~Timepoint)+
+  facet_wrap(Timepoint ~ Donor, scales = "free_x")+
 
   theme(axis.text.x = element_text(angle = 25))+
-  ggtitle("RANTES data")
-
-```
-
-```{r RANTES alternate, echo=FALSE, warning=FALSE, message= FALSE}
-
-ggplot(data = RANTESexplantDat, aes(x = Timepoint, y = Calc..Concentration))+
-
-  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5 )+
-
-  stat_summary(aes(group = Donor, color = Donor),fun.y = mean, geom = "point", size = 2)+
-
-  #stat_summary(aes(group = Donor, color = Donor), fun.y =mean, geom = "line")+
-
-  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+
-
-  scale_y_log10()+
-
-  labs( y = "Concentration")+
-
-  facet_wrap(~Virus)+
   ggtitle("RANTES data")
@@
 
-  facet_wrap(~Timepoint)+
+  facet_wrap(Timepoint ~ Donor, scales = "free_x")+
 
@@
 ```
-
-```{r IL1RA alternate, echo=FALSE, warning=FALSE, message= FALSE}
-
-ggplot(data = IL1RAexplantDat, aes(x = Timepoint, y = Calc..Concentration))+
-
-  geom_point(aes(color = Donor, shape = Detection.Range ),size = 1, alpha = 0.5 )+
-
-  stat_summary(aes(group = Donor, color = Donor),fun.y = mean, geom = "point", size = 2)+
-
-  #stat_summary(aes(group = Donor, color = Donor), fun.y =mean, geom = "line")+
-
-  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+
-
-  scale_y_log10()+
-
-  labs( y = "Concentration")+
-
-  facet_wrap(~Virus)+
-  ggtitle("IL1RA data")
-```

### 2016-07-19 30June16_plate_3_explant_sup.Rmd
updated plots

@@
   filter(Sample != "Diluted Stock Calibrator" & Sample != "Blended Working Stock")
+
 ```
@@
 ```{r standard curves, echo=FALSE, warning=FALSE, message= FALSE}
+#according to MSD instructions, the lower detection limit for the standards (in signal units) is the signal for the lower calibrator + (statistical factor * multiplier)
 
-# standardCurve<- function (Calc.Top, Calc.Bottom, Calc.HillSlope, Calc.Midpoint, Concentration){
+# upper limit is calculated this way: upper plateau of standard curve - (multiplier x statistical factor). The upper plateau value is in the "Algorithm Parameter Calc Top" column.
+
+#the statistical factor used can be found in the "Detection limits bottom of range type" column and the multiplier can  be found in the "Detection limits bottom of range parameter" column
+
+#for the 4PL curve, std deviation is the statistical factor used for both limits.
+
+# LowerLimits <-standards %>%
+#   group_by(Assay)%>%
+#   filter(Sample == "STD-08")%>%
+#   summarize(LowerLimit = unique(Mean + (Std..Deviation * Detection.Limits..Bottom.of.Range.Parameter)))
 # 
-# Calc.Top + (Calc.Bottom - Calc.Top/(1+(Concentration/Calc.Midpoint)^Calc.HillSlope))
-# }
+# UpperLimits<- standards %>%
+#   group_by(Assay)%>%
+#   summarise(UpperLimit = unique(Algorithm.Parameter..Calc..Top - (Std..Deviation * Detection.Limits..Top.of.Range.Parameter)))
 # 
+# standards<-merge(standards,LowerLimits, by = "Assay")
 # 
-# MIGcurveData <- standards%>%
-#   filter(Assay == "MIG") %>%
-#   mutate(UpperLimit = Algorithm.Parameter..Calc..Top-( Detection.Limits..Top.of.Range.Parameter * Std..Deviation ))%>%
-#   mutate(LowerLimit = min(Concentration)+(2.5 * Std..Deviation ))
+# standards<- merge(standards, UpperLimits, by = "Assay")
 # 
@@
 # 
-# MIGcalcCurve <- standardCurve(MIGcurveData$Algorithm.Parameter..Calc..Top,MIGcurveData$Algorithm.Parameter..Calc..Bottom, MIGcurveData$Algorithm.Parameter..Calc..HillSlope, MIGcurveData$Algorithm.Parameter..Calc..MidPoint, MIGcurveData$Concentration)
+# MIGstandards<- filter(standards, Assay == "MIG")
 # 
-# 
-# ggplot(MIGcurveData, aes(x = Concentration, y = Signal))+
+# MIGstandards%>%
+#   ggplot(aes(x = Concentration, y = Signal))+
 #   geom_point()+
@@
 #   scale_x_log10()+
-#   geom_line(aes(y = MIGcalcCurve))+
-#   geom_hline(yintercept = MIGcurveData$UpperLimit)
+#   geom_hline(yintercept = MIGstandards$UpperLimit[1])+
+#   geom_hline(yintercept = MIGstandards$LowerLimit[1])+
+#   ggtitle("MIG")
+#   
+# 
+# RANTESstandards<- filter(standards, Assay == "RANTES")
+# 
+# RANTESstandards%>%
+#   filter(Assay == "RANTES")%>%
+#   ggplot(aes(x = Concentration, y = Signal))+
+#   geom_point()+
+#   scale_y_log10()+
+#   scale_x_log10()+
+#   geom_hline(yintercept = RANTESstandards$UpperLimit[1])+
+#   geom_hline(yintercept = RANTESstandards$LowerLimit[1])+
+#   ggtitle("RANTES")
+# 
+# IL1RAstandards<-filter(standards, Assay == "IL1-RA")
+# 
+# IL1RAstandards%>%
+#   filter(Assay == "IL1-RA")%>%
+#   ggplot(aes(x = Concentration, y = Signal))+
+#   geom_point()+
+#   scale_y_log10()+
+#   scale_x_log10()+
+#   geom_hline(yintercept = IL1RAstandards$UpperLimit[1])+
+#   geom_hline(yintercept = IL1RAstandards$LowerLimit[1])+
+#   ggtitle("IL1-RA")
 
@@
 
-```{r everything, echo=FALSE, warning=FALSE, message= FALSE}
+```{r all analytes, echo=FALSE, warning=FALSE, message= FALSE}
 
@@
 ```
+
+When we analyzed the explant microarray data, we saw the most obvious differences in gene expression were in V2T3 for donors 323,317,324,318.
+The only sample that was in this subset that we ran on this plate was 323.
+
+```{r subset, echo=FALSE, warning=FALSE, message= FALSE}
+subset <- explantDat %>%
+  filter(Donor!="319" & Donor != "319" & Donor != "327" & Donor!= "326", Timepoint == "T3")
+
+ggplot(subset, aes(x = Virus, y = Calc..Concentration))+
+
+  geom_point(aes(color = Assay, shape = Detection.Range ),size = 1.5)+
+
+  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Fit Curve Range" = 1, "In Detection Range" = 16))+
+
+  scale_y_log10()+
+
+  labs( y = "Concentration")+
+
+  facet_wrap(Timepoint~Donor, scales = "free_x")+
+
+  theme(axis.text.x = element_text(angle = 25, size = 9))
+
+```
+
+Data separated by analyte:
 

### 2016-08-15 MSD_panels_1_2_3_explant_sup_analysis.Rmd
Analysis of panels 2 and 3 only

---
title: "MSD panels 1, 2 and 3 Explant Sup"
author: "Claire Levy"
date: "August 12, 2016"
output: github_document
---

## Experiment overview

These data are from running MSD analysis of supernatents from selected vaginal explant samples. The explants were infected with HSV2 strains 186, SD90 or Mock infected by Lamar Fleming and Gabriella Fenkart.

We used 3 different MSD panels containing a total of 20 analytes analyzed for 46 samples(see caveat). Bsaed on results from running test samples, we decided to dilute all the supernants 1:5 with MSD's diluent 43 before running the plates. All samples (except 1, see caveats) were run in duplicate.

Caveats:
Due to limited sample volume, we had to make the following omissions:

Panel 2: Only 1 rep for sample 324-T2-V1

Panel 3: Did not run samples 324-T2-V1 or 324-T2-V2.

Due to a possible pipetting error, we were not able to use the data from the standards that we ran for panel 1. Instead, we used the standards from a previous run of the same panel to calculate concentrations for the experimental samples(see explant_sup_README for details).

```{r data read-in, echo=FALSE, message = FALSE, warning = FALSE}

library(dplyr)
library(ggplot2)
library(pander)
library(reshape2)
library(stringr)

#reading in data from the panel 1

#remember that this data table contains data from the panel 1-Test from 28Mar16 (Plate_2BF6AAN447). I am only going to use the Standards data from that plate and not the standards from either of the plates for panel 1 run 10Aug16.

# HOWEVER: When we did panel 1-TEST, the kit that we used was missing the Ab for MIP1B, so we didn't include that one the plate, so we don't have any standard data for that analyte from this plate. So, we will need to use the standard data from the 10Aug16 run of panel 1 for that analyte, or, not use it.

#annoyingly, for Plate_2BF6AAN447, the standards are not named "STD-01" etc in the sample names column but rather "S001". I'm going to change them to match the other panels.

panel_1 <- read.csv("../raw_data/10Aug16_explant_sup_panel_1_plate_1_and_2.csv", skip = 1)

TestStdsToUse <- panel_1 %>%
  filter(Sample.Group == "Standards")%>%
  filter(Plate.Name == "panel_1_TEST")

#fixing the sample names!
TestStdsToUse$Sample<-str_replace(TestStdsToUse$Sample, "S0","STD-")

#filtering out just the standards for MIP1B from the 11Aug16 panel 1 plates (don't have data for that analyte in the test panel). The greek letters got all messed up since they went thru excel, but I know which is which based on the Spot (in the well) that they are associated with.

StdsToUse<- panel_1 %>%
  filter(Sample.Group == "Standards", Spot == "4")

#now filter out all the standards from the panel 1 data and add back in just the standards I want (TestStdsToUse and StdsToUse)

panel_1NoStds<-panel_1 %>%
  filter(Sample.Group != "Standards")

panel_1_Stds_corrected <- rbind(panel_1NoStds,StdsToUse,TestStdsToUse )

# Data from panels 2 and 3
panel_2<-read.csv("../raw_data/10Aug16_explant_sup_panel_2_plate_1_and_2.csv", skip = 1)

panel_3<-read.csv("../raw_data/11Aug16_explant_sup_panel_3_plate_1_and_2.csv", skip = 1)

#Combining the data from all 3 panels

allPanels<- rbind(panel_1_Stds_corrected, panel_2,panel_3)

```

```{r fixing panel 1 standards, echo=FALSE, message = FALSE, warning = FALSE}

#just the standards from all the panels

##SANITY CHECK
# panel 1: 8std x 2reps x 1analytes x 2 plates = 32
# panel 1 TEST: 8std x 2reps x 8analyte x 1plate = 128
# panel 2: 8std x 2reps x 8analytes x 2 plates = 256
# panel 3: 8std x 2reps x 3analytes x 2 plates = 96

# 32+128+256+96=512

allPanelsStds<-allPanels %>%
  filter(Sample.Group == "Standards")%>%
  group_by(Plate.Name)
```

## Checking Standard Recovery and Curves

Here are plots of the % recovery (extrapolated concentration/ known concentration x 100) for the standards that were either within or above the detection range, with lines at the "good recovery" limits of 80 and 120% recovery.

The R^2 values for goodness-of-fit for the standard curves were all >0.99
```{r panel 2 and 3 standards, echo=FALSE, message = FALSE, warning = FALSE}

#I want to ignore panel 1 data right now since we might get a better set of standards to use with it.

panel_2_and_3<- rbind(panel_3, panel_2)

panel_2_and_3Stds<-filter(panel_2_and_3, Sample.Group == "Standards")

#which standards are out of det. range?
standardsOutOfRange<-panel_2_and_3Stds %>%
  filter(Detection.Range != "In Detection Range")%>%
  arrange(Plate.Name,Sample, Assay, Detection.Range)%>%
  select(Plate.Name,Sample, Assay, Detection.Range)

stdFit<- panel_2_and_3Stds %>% 
  arrange(Sample,Assay)%>%
  filter(Detection.Range != "Below Fit Curve Range")%>%
  select(Sample, Assay,Detection.Range, X..Recovery,Fit.Statistic..RSquared, Concentration, Calc..Concentration)

ggplot(stdFit, aes(x = Sample, y = X..Recovery))+
  geom_point(aes(color = Sample, shape = Detection.Range ),size = 2, alpha = 0.5)+
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "In Detection Range" = 16, "Below Detection Range" = 0))+
  facet_wrap(~Assay)+
  theme(axis.text.x = element_text(angle=25))+
  scale_y_continuous(limits = c(0, 175))+
  geom_abline(slope = 0, intercept = c(80, 120))+
  labs(y = "Percent Recovery")+
  theme(axis.text.x = element_text(size = rel(0.8)))

#checking the R squared goodness-of-fit for the standard curves
rSquared<-stdFit%>%
  group_by(Assay)%>%
  summarize(R_Squared = mean(Fit.Statistic..RSquared))

```

```{r clean data, echo=FALSE, message = FALSE, warning = FALSE}

#Making a df that contains just the non-standards

panel_2_and_3_noStds <- panel_2_and_3 %>%
  filter(Sample.Group!="Standards")

# I want to be able to group the data by timepoint and donor and virus so I need to separate the sample name and make new columns identifying them that way

donorTimeVirus<-colsplit(panel_2_and_3_noStds$Sample,"-", c("Donor","Timepoint","Virus") )

#bind the split col back to panel_2_and_3

panel_2_and_3_noStds<-cbind(donorTimeVirus, panel_2_and_3_noStds)

#Make things factors so they plot nicely
panel_2_and_3_noStds$Donor <- as.factor(panel_2_and_3_noStds$Donor)

panel_2_and_3_noStds$Calc..Concentration <-as.numeric(panel_2_and_3_noStds$Calc..Concentration)

panel_2_and_3_noStds$Virus <- as.factor(panel_2_and_3_noStds$Virus)

panel_2_and_3_noStds$Virus<-factor(panel_2_and_3_noStds$Virus, levels = c("M", "V1","V2"))

panel_2_and_3_noStds$Timepoint<-as.factor(panel_2_and_3_noStds$Timepoint)
```

```{r checking reps, echo=FALSE, message = FALSE, warning = FALSE}

#checking to make sure that only the samples that SHOULD have 1 rep, show up as having just 1 rep

replicateSummary <- panel_2_and_3_noStds %>%
group_by(Plate.Name,Sample, Assay)%>%
summarize(reps = n())

not2<-filter(replicateSummary,reps!="2")%>%
  arrange(Sample, Assay)

```

```{r avg reps, echo=FALSE, message = FALSE, warning = FALSE}

#some of the samples had one rep that was below detection, so I am excluding those so I will be able to use the remaining rep as the "average" for that sample.

#I am also excluding samples that are "Below Detection Range" because this means that the signal was "statistically indistinguishable from the background or top of the curve"" and ..."The concentrations calculated are inherently unreliable."

#Which were below detection range?
panel_2_and_3_noStds_Below <- panel_2_and_3_noStds %>%
  filter(Detection.Range == "Below Fit Curve Range" & Detection.Range != "Below Detection Range")

panel_2_and_3_noStds_noBelow <- panel_2_and_3_noStds %>%
  filter(Detection.Range != "Below Fit Curve Range" & Detection.Range != "Below Detection Range")

#averaging the replicates
avgConcpanel_2_and_3_noStds_noBelow <- panel_2_and_3_noStds_noBelow %>%
group_by(Sample, Assay, Donor, Timepoint,Virus)%>%
summarize(avgConc = mean(Calc..Concentration))%>%
  ungroup()

#Looking at the number of samples in each sample "set"(Same Donor and Timepoint). Some are missing V1 or V2. or both!! I don't want the ones with just one sample in the set.

numberInSets<-avgConcpanel_2_and_3_noStds_noBelow %>%
         group_by(Donor,Timepoint,Assay)%>%
  select(Sample,Donor, Timepoint,Assay)%>%
  summarize(numInSet = n())

#now I'll merge the avgConc df with this numberInSets info created above so I can then filter out samples that have <2 samples in the set

avgConcpanel_2_and_3_noStds_noBelow <- merge(avgConcpanel_2_and_3_noStds_noBelow, numberInSets, by = c("Donor", "Timepoint", "Assay"))

#now exclude the samples where there not a "full set", i.e. 1 sample only in the set

avgConcpanel_2_and_3_noStds_noBelow_fullSet <- avgConcpanel_2_and_3_noStds_noBelow %>%
  filter(numInSet != 1)

```

## Plots of analyte concentrations

## Caveats
Samples omitted from the plots include:

* Concentrations below the fit curve range. Values cannot be extrapolated when the data was *below* the fit curve range. However,the MSD software can extrapolate values that are *above* the fit curve range if the curve is linear at the top.

* Samples for which there was only data for a Mock sample and neither of the virus conditions.

Also note:
Some of the samples had one rep that was below detection and another that was above. For these samples I used the remaining rep as the "average" for that sample.

```{r  samples plots, echo=FALSE, message = FALSE, warning = FALSE}

#make a list of the analytes
analyteList<-as.list(levels(avgConcpanel_2_and_3_noStds_noBelow$Assay))

#write a function to make a plot for each analyte

plotAnalytes <- function (Analyte){
   avgConcpanel_2_and_3_noStds_noBelow_fullSet %>%
     filter(Assay == Analyte)%>%

     ggplot( aes(x = Virus, y = avgConc))+

   geom_point(aes(color = Donor ),size = 1.5)+
  geom_line(aes(group = Donor, color = Donor)) +

   scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Detection Range" = 1, "In Detection Range" = 16))+

   scale_y_log10()+

   labs( y = "Concentration")+

  facet_wrap(~Timepoint, scales = "free_x")+

   theme(axis.text.x = element_text(angle = 25, size = 9))+
     ggtitle(Analyte)
}

lapply(analyteList, FUN = plotAnalytes)

pander(panel_2_and_3_noStds %>%
  select(Sample, Assay, Calc..Concentration, Detection.Range))

### 2016-08-19 MSD_panels_1_2_3_explant_sup_analysis.Rmd
Updated code to include panel 1 data, added in a bunch of sanity checks.

@@
 
-Caveats:
+## Caveats
+
 Due to limited sample volume, we had to make the following omissions:
@@
 Due to a possible pipetting error, we were not able to use the data from the standards that we ran for panel 1.
-Instead, we used the standards from a previous run of the same panel to calculate concentrations for the experimental samples(see explant_sup_README for details).
+Instead, we used the standards from a different run of the same panel to calculate concentrations for the experimental samples(see explant_sup_README for details).
 
@@
 
-#remember that this data table contains data from the panel 1-Test from 28Mar16 (Plate_2BF6AAN447).
-I am only going to use the Standards data from that plate and not the standards from either of the plates for panel 1 run 10Aug16.
+#remember that this data table contains data from the 17Aug16_HVE_sup_panel_1_plate_1.
+I am only going to use the Standards data from that plate and not the standards from either of the plates for panel 1 run 10Aug16 since we had what appears to be a pipetting error (2nd replicate is more concentration for all wells).
 
-# HOWEVER: When we did panel 1-TEST, the kit that we used was missing the Ab for MIP1B, so we didn't include that one the plate, so we don't have any standard data for that analyte from this plate.
-So, we will need to use the standard data from the 10Aug16 run of panel 1 for that analyte, or, not use it.
+explant_panel_1_and_HVE_panel_1_plate_1 <- read.csv("../raw_data/10Aug16_explant_sup_panel_1_plate_1_and_2.csv", skip = 1)
 
-#annoyingly, for Plate_2BF6AAN447, the standards are not named "STD-01" etc in the sample names column but rather "S001". I'm going to change them to match the other panels.
+#Filter to ONLY keep the standards from the plate called "HVE_sup_panel_1_plate_1"
 
-panel_1 <- read.csv("../raw_data/10Aug16_explant_sup_panel_1_plate_1_and_2.csv", skip = 1)
+StdsToUse_explant_panel_1 <- explant_panel_1_and_HVE_panel_1_plate_1 %>%
+  filter(Sample.Group == "Standards")%>%
+  filter(Plate.Name == "HVE_sup_panel_1_plate_1")
 
-TestStdsToUse <- panel_1 %>%
-  filter(Sample.Group == "Standards")%>%
-  filter(Plate.Name == "panel_1_TEST")
+#Since I told Discovery Workbench software that I wanted to use the standards from the 17Aug16_HVE_sup_panel_1_plate_1 as "Global Standards" (i.e.
+use them to calculate concentrations for all plates in the experiment)it seems to have provided me with 3 copies of those samples in the data table (1 for each plate I guess).
+I only need data from the unique entries, so I will filter for just unique rows in this df to exclude copies.
 
-#fixing the sample names!
-TestStdsToUse$Sample<-str_replace(TestStdsToUse$Sample, "S0","STD-")
+# should have 8 stds x 9 analytes x 2 reps = 144 obs
+StdsToUse_explant_panel_1<-unique(StdsToUse_explant_panel_1)
 
-#filtering out just the standards for MIP1B from the 11Aug16 panel 1 plates (don't have data for that analyte in the test panel).
-The greek letters got all messed up since they went thru excel, but I know which is which based on the Spot (in the well) that they are associated with.
+#Sanity check: there should only be 2 entries for each unique instance of Sample (STD-0x) and assay. 
+checkStdReps<-StdsToUse_explant_panel_1 %>%
+         group_by(Sample,Assay)%>%
+         summarise(n())
 
-StdsToUse<- panel_1 %>%
-  filter(Sample.Group == "Standards", Spot == "4")
+#Remove the 17Aug16_HVE_sup_panel_1_plate_1 non-STD data from the df and leave ONLY explant panel 1 data. Remaining should be:
+#94 wells from plate 1 + 30 wells from plate 2 x 9 analytes = 1116 observations
 
-#now filter out all the standards from the panel 1 data and add back in just the standards I want (TestStdsToUse and StdsToUse)
+explant_panel_1 <-explant_panel_1_and_HVE_panel_1_plate_1 %>%
+  filter(Plate.Name != "HVE_sup_panel_1_plate_1")
 
-panel_1NoStds<-panel_1 %>%
+#now remove the unwanted Std samples from the explant data. There should be 32 Std wells x 9 analytes = 288
+# 1116 total - 288 = 828 observations
+
+No_Stds_explant_panel_1 <-explant_panel_1 %>%
   filter(Sample.Group != "Standards")
 
-panel_1_Stds_corrected <- rbind(panel_1NoStds,StdsToUse,TestStdsToUse )
+#now add in the unique Stds from 17Aug16_HVE_sup_panel_1_plate_1
+
+corrected_Stds_explant_panel_1 <- rbind(No_Stds_explant_panel_1,StdsToUse_explant_panel_1)
 
 # Data from panels 2 and 3
-panel_2<-read.csv("../raw_data/10Aug16_explant_sup_panel_2_plate_1_and_2.csv", skip = 1)
 
-panel_3<-read.csv("../raw_data/11Aug16_explant_sup_panel_3_plate_1_and_2.csv", skip = 1)
+#panel 2 should have 123 wells (left out one rep) x 8 analytes = 984
+explant_panel_2<-read.csv("../raw_data/10Aug16_explant_sup_panel_2_plate_1_and_2.csv", skip = 1)
 
-#Combining the data from all 3 panels
+#panel 3: 120 wells (left out 2 samples, 4 reps) x 8 analytes =360
 
-allPanels<- rbind(panel_1_Stds_corrected, panel_2,panel_3)
+explant_panel_3<-read.csv("../raw_data/11Aug16_explant_sup_panel_3_plate_1_and_2.csv", skip = 1)
 
-```
+#Combining the data from all 3 panels: 972 +984 + 360= 2316
 
-```{r fixing panel 1 standards, echo=FALSE, message = FALSE, warning = FALSE}
+explant_all_panels<- rbind(corrected_Stds_explant_panel_1, explant_panel_2,explant_panel_3)
 
-#just the standards from all the panels
-
-##SANITY CHECK
-# panel 1: 8std x 2reps x 1analytes x 2 plates = 32
-# panel 1 TEST: 8std x 2reps x 8analyte x 1plate = 128
-# panel 2: 8std x 2reps x 8analytes x 2 plates = 256
-# panel 3: 8std x 2reps x 3analytes x 2 plates = 96
-
-# 32+128+256+96=512
-
-allPanelsStds<-allPanels %>%
-  filter(Sample.Group == "Standards")%>%
-  group_by(Plate.Name)
 ```
@@
 
-The R^2 values for goodness-of-fit for the standard curves were all >0.99
+The R^2^ values for goodness-of-fit for the standard curves were all >0.99
+
 ```{r panel 2 and 3 standards, echo=FALSE, message = FALSE, warning = FALSE}
 
-#I want to ignore panel 1 data right now since we might get a better set of standards to use with it.
-
-panel_2_and_3<- rbind(panel_3, panel_2)
-
-panel_2_and_3Stds<-filter(panel_2_and_3, Sample.Group == "Standards")
+explant_stds_all_panels<-filter(explant_all_panels, Sample.Group == "Standards")
 
 #which standards are out of det. range?
-standardsOutOfRange<-panel_2_and_3Stds %>%
+standardsOutOfRange<-explant_stds_all_panels %>%
   filter(Detection.Range != "In Detection Range")%>%
-  arrange(Plate.Name,Sample, Assay, Detection.Range)%>%
-  select(Plate.Name,Sample, Assay, Detection.Range)
+  select(Plate.Name,Sample, Assay, Detection.Range, X..Recovery, Calc..Concentration)
 
-stdFit<- panel_2_and_3Stds %>% 
+#looking at which samples have STDs below the fit curve or below the detection range (i.e. you can't get numbers for % recovery) 
+explant_stds_all_panels_below<- explant_stds_all_panels%>%
+  filter(Detection.Range == "Below Fit Curve Range" | Detection.Range == "Below Detection Range") %>%
+  group_by(Sample)%>%
+  summarise("Number of Samples Below Fit Curve Range/Detection Range" = n())
+
+#The samples that actually give me #'s for percent recovery ( i.e. all but the "below fit curve range" and "below detection range" ones)
+
+explant_stds_all_panels_percent_recov<- explant_stds_all_panels %>% 
   arrange(Sample,Assay)%>%
-  filter(Detection.Range != "Below Fit Curve Range")%>%
-  select(Sample, Assay,Detection.Range, X..Recovery,Fit.Statistic..RSquared, Concentration, Calc..Concentration)
+  filter(Detection.Range != "Below Fit Curve Range" & Detection.Range!= "Below Detection Range")%>%
+  select(Plate.Name,Sample, Assay,Detection.Range, X..Recovery,Fit.Statistic..RSquared, Concentration, Calc..Concentration)
 
-ggplot(stdFit, aes(x = Sample, y = X..Recovery))+
-  geom_point(aes(color = Sample, shape = Detection.Range ),size = 2, alpha = 0.5)+
-  scale_shape_manual(values = c("Above Fit Curve Range" = 1, "In Detection Range" = 16, "Below Detection Range" = 0))+
+#plotting percent recovery
+
+ggplot(explant_stds_all_panels_percent_recov,aes( x = Sample, y = X..Recovery)) +
+  geom_point(aes(shape = Detection.Range ),
+             size = 2, alpha = 0.5)+
+  scale_shape_manual(values = c("Above Fit Curve Range" = 1, 
+                                "In Detection Range" = 16, 
+                                "Below Detection Range" = 0))+
   facet_wrap(~Assay)+
   theme(axis.text.x = element_text(angle=25))+
-  scale_y_continuous(limits = c(0, 175))+
+  scale_y_continuous(limits = c(0, 250))+
   geom_abline(slope = 0, intercept = c(80, 120))+
@@
 #checking the R squared goodness-of-fit for the standard curves
-rSquared<-stdFit%>%
+rSquared<-explant_stds_all_panels_percent_recov%>%
   group_by(Assay)%>%
@@
 
-```{r clean data, echo=FALSE, message = FALSE, warning = FALSE}
+## Caveats about Standards
+
+* Standards with values below the detection range (<2.5x the standard deviation of the lowest standard) are give unreliable concentrations, so I did not include those in the plots.
+
+* When standards are below the curve fit (outside the range of the standard curve), the software does not calculate concentrations so you can't get a percent recovery.
+
+* Most of the Standards that fell into either of these categories were from STD-07 or STD-08 except for two that were STD-06:
+
+`r pander(explant_stds_all_panels_below)r`
+
+```{r break up sample name, echo=FALSE, message = FALSE, warning = FALSE}
 
@@
 
-panel_2_and_3_noStds <- panel_2_and_3 %>%
+explant_all_panels_noStds <- explant_all_panels %>%
   filter(Sample.Group!="Standards")
 
-# I want to be able to group the data by timepoint and donor and virus so I need to separate the sample name and make new columns identifying them that way
+# I want to be able to group the data by timepoint and donor and virus so I need to break up components of the the sample name and make new columns with just the Donor, Timepoint and Virus
 
-donorTimeVirus<-colsplit(panel_2_and_3_noStds$Sample,"-", c("Donor","Timepoint","Virus") )
+#Split samples names on the "-"
+donorTimeVirus<-colsplit(explant_all_panels_noStds$Sample,"-", c("Donor","Timepoint","Virus") )
 
-#bind the split col back to panel_2_and_3
+#bind the split col back to allPanels
 
-panel_2_and_3_noStds<-cbind(donorTimeVirus, panel_2_and_3_noStds)
+explant_all_panels_noStds<-cbind(donorTimeVirus, explant_all_panels_noStds)
 
 #Make things factors so they plot nicely
-panel_2_and_3_noStds$Donor <- as.factor(panel_2_and_3_noStds$Donor)
+explant_all_panels_noStds$Donor <- as.factor(explant_all_panels_noStds$Donor)
 
-panel_2_and_3_noStds$Calc..Concentration <-as.numeric(panel_2_and_3_noStds$Calc..Concentration)
+explant_all_panels_noStds$Calc..Concentration <-as.numeric(explant_all_panels_noStds$Calc..Concentration)
 
-panel_2_and_3_noStds$Virus <- as.factor(panel_2_and_3_noStds$Virus)
+explant_all_panels_noStds$Virus <- as.factor(explant_all_panels_noStds$Virus)
 
-panel_2_and_3_noStds$Virus<-factor(panel_2_and_3_noStds$Virus, levels = c("M", "V1","V2"))
+explant_all_panels_noStds$Virus<-factor(explant_all_panels_noStds$Virus, levels = c("M", "V1","V2"))
 
-panel_2_and_3_noStds$Timepoint<-as.factor(panel_2_and_3_noStds$Timepoint)
+explant_all_panels_noStds$Timepoint<-as.factor(explant_all_panels_noStds$Timepoint)
 ```
@@
 
-#checking to make sure that only the samples that SHOULD have 1 rep, show up as having just 1 rep
+#checking to make sure that only the samples that SHOULD have 1 rep, show up as having just 1 rep (sample 324-T2-V1 where I ran out of sample before I could do the 2nd rep)
 
-replicateSummary <- panel_2_and_3_noStds %>%
+replicateSummary <- explant_all_panels_noStds %>%
 group_by(Plate.Name,Sample, Assay)%>%
@@
   arrange(Sample, Assay)
+
+```
+
+## Explant Supernatents: Detection Range
+
+```{r checking detection range, echo=FALSE, message = FALSE, warning = FALSE}
+
+#What number of samples are in each category of detection?
+explant_all_panels_noStds_Below <- explant_all_panels_noStds%>%
+ group_by(Assay, Detection.Range)%>%
+  summarise(Number_of_Samples = n())
+
+#Let's plot that
+
+ggplot(explant_all_panels_noStds_Below, aes(x = Assay, y=Number_of_Samples))+
+         geom_point(aes(color = Detection.Range),size = 2)+
+  theme(axis.text.x = element_text(angle=25))+
+  ggtitle("Number of Samples in each Detection Range per Analyte")
 
@@
 
-#Which were below detection range?
-panel_2_and_3_noStds_Below <- panel_2_and_3_noStds %>%
-  filter(Detection.Range == "Below Fit Curve Range" & Detection.Range != "Below Detection Range")
-
-panel_2_and_3_noStds_noBelow <- panel_2_and_3_noStds %>%
+explant_all_panels_noStds_noBelow <- explant_all_panels_noStds %>%
   filter(Detection.Range != "Below Fit Curve Range" & Detection.Range != "Below Detection Range")
@@
 #averaging the replicates
-avgConcpanel_2_and_3_noStds_noBelow <- panel_2_and_3_noStds_noBelow %>%
+avgConc_explant_all_panels_noStds_noBelow <- explant_all_panels_noStds_noBelow %>%
 group_by(Sample, Assay, Donor, Timepoint,Virus)%>%
@@
 
-numberInSets<-avgConcpanel_2_and_3_noStds_noBelow %>%
+numberInSets<-avgConc_explant_all_panels_noStds_noBelow %>%
          group_by(Donor,Timepoint,Assay)%>%
@@
 
-avgConcpanel_2_and_3_noStds_noBelow <- merge(avgConcpanel_2_and_3_noStds_noBelow, numberInSets, by = c("Donor", "Timepoint", "Assay"))
+avgConc_explant_all_panels_noStds_noBelow <- merge(avgConc_explant_all_panels_noStds_noBelow, numberInSets, by = c("Donor", "Timepoint", "Assay"))
 
@@
 
-avgConcpanel_2_and_3_noStds_noBelow_fullSet <- avgConcpanel_2_and_3_noStds_noBelow %>%
+avgConc_explant_all_panels_noStds_noBelow_fullSet <- avgConc_explant_all_panels_noStds_noBelow %>%
   filter(numInSet != 1)
@@
 ## Plots of analyte concentrations
+
+Black points show the means.
 
@@
 Some of the samples had one rep that was below detection and another that was above. For these samples I used the remaining rep as the "average" for that sample.
+```{r correcting analyte names, echo=FALSE, message = FALSE, warning = FALSE}
+
+#excel messes up the greek letters of the analytes so I'm fixing them here.
+
+#unicode greek lower case alpha = U03B1
+
+```
 
@@
 #make a list of the analytes
-analyteList<-as.list(levels(avgConcpanel_2_and_3_noStds_noBelow$Assay))
+analyteList<-as.list(levels(avgConc_explant_all_panels_noStds_noBelow$Assay))
 
@@
 plotAnalytes <- function (Analyte){
-   avgConcpanel_2_and_3_noStds_noBelow_fullSet %>%
+   avgConc_explant_all_panels_noStds_noBelow_fullSet %>%
      filter(Assay == Analyte)%>%
 
-     ggplot( aes(x = Virus, y = avgConc))+
+     ggplot( aes(x = Timepoint, y = avgConc))+
 
-   geom_point(aes(color = Donor ),size = 1.5)+
-  geom_line(aes(group = Donor, color = Donor)) +
+   geom_point(aes(color = Donor ),size = 1.5, alpha = 0.5)+
+  geom_line(aes(group = Donor, color = Donor), alpha = 0.5)+
+    stat_summary(fun.y = mean, geom = "point")+
+    stat_summary(fun.y = mean, geom = "line",aes(group = 1))+
 
@@
 
-  facet_wrap(~Timepoint, scales = "free_x")+
+  facet_wrap(~Virus, scales = "free_x")+
 
@@
 lapply(analyteList, FUN = plotAnalytes)
-
-pander(panel_2_and_3_noStds %>%
-  select(Sample, Assay, Calc..Concentration, Detection.Range))
+```

### 2016-08-22 MSD_panels_1_2_3_explant_sup_analysis.Rmd
fixed greek letters, changed detection range plot

@@
 
-We used 3 different MSD panels containing a total of 20 analytes analyzed for 46 samples(see caveat).
-Bsaed on results from running test samples, we decided to dilute all the supernants 1:5 with MSD's diluent 43 before running the plates.
+We used 3 different MSD panels containing a total of 20 analytes analyzed for 46 samples (see caveat).
+Based on results from running test samples, we decided to dilute all the supernants 1:5 with MSD's diluent 43 before running the plates.
 All samples (except 1, see caveats) were run in duplicate.
@@
 Due to a possible pipetting error, we were not able to use the data from the standards that we ran for panel 1.
-Instead, we used the standards from a different run of the same panel to calculate concentrations for the experimental samples(see explant_sup_README for details).
+Instead, we used the standards from a different run of the same panel to calculate concentrations for the experimental samples (see explant_sup_README for details).
 
@@
 
+#fixing the greek letters!!
+explant_all_panels$Assay <-recode(explant_all_panels$Assay,
+                                      "IFN-α2a" = paste("IFN-\U03B1","2a", sep = ""),
+                                      "IL-1α" = "IL-1\U03B1",
+                                      "MIP-1α"="MIP-1\U03B1",
+                                      "TNF-α" = "TNF-\U03B1",
+                                      "MIP-3α" = "MIP-3\U03B1",
+                                      "MIP-1β" = "MIP-1\U03B2",
+                                      "IL-1β" = "IL-1\U03B2",
+                                      "IFN-γ" = "IFNg")
+
 ```
@@
 
-```{r panel 2 and 3 standards, echo=FALSE, message = FALSE, warning = FALSE}
+```{r standards, echo=FALSE, message = FALSE, warning = FALSE}
 
@@
 
-* Standards with values below the detection range (<2.5x the standard deviation of the lowest standard) are give unreliable concentrations, so I did not include those in the plots.
+* Standards with values *below the detection range* (<2.5x the standard deviation of the lowest standard) give unreliable concentrations, so I did not include those in the plots.
 
-* When standards are below the curve fit (outside the range of the standard curve), the software does not calculate concentrations so you can't get a percent recovery.
+* When standards are *below the curve fit* (outside the range of the standard curve), the software does not calculate concentrations so you can't get a percent recovery.
 
@@
 
-`r pander(explant_stds_all_panels_below)r`
+`r pander(explant_stds_all_panels_below)`
 
@@
 explant_all_panels_noStds_Below <- explant_all_panels_noStds%>%
- group_by(Assay, Detection.Range)%>%
+  mutate(Panel = ifelse(str_detect(Plate.Name, "panel_1"),"Panel 1",ifelse(str_detect(Plate.Name,"panel_2"),"Panel 2","Panel 3")))%>%
+ group_by(Panel,Assay, Detection.Range)%>%
   summarise(Number_of_Samples = n())
@@
   theme(axis.text.x = element_text(angle=25))+
+  labs(y = "Number of Samples", x = "Analyte")+
+  facet_wrap(~Panel, scales = "free")+
   ggtitle("Number of Samples in each Detection Range per Analyte")
@@
 Also note:
-Some of the samples had one rep that was below detection and another that was above. For these samples I used the remaining rep as the "average" for that sample.
-```{r correcting analyte names, echo=FALSE, message = FALSE, warning = FALSE}
-
-#excel messes up the greek letters of the analytes so I'm fixing them here.
-
-#unicode greek lower case alpha = U03B1
-
-```
+Some of the samples had one rep that was below detection and another that was in range. For these samples I used the remaining "in range" rep as the "average" for that sample.
 
@@
 #make a list of the analytes
-analyteList<-as.list(levels(avgConc_explant_all_panels_noStds_noBelow$Assay))
+
+#I'm using a unique vector here instead of the levels of Assay because I don't want to include IFNa2a OR Il-7.
+There weren't any data points in the detection range for those analytes (so I don't want to try to make a plot of it)
+
+analyteList<-as.list(unique(avgConc_explant_all_panels_noStds_noBelow$Assay))
 
@@
 lapply(analyteList, FUN = plotAnalytes)
+
 ```

### 2016-08-23 MSD_panels_1_2_3_explant_sup_analysis.Rmd
changed dfs to just have columns of interest and adjusted sup plots.

@@
 
-These data are from running MSD analysis of supernatents from selected vaginal explant samples.
-The explants were infected with HSV2 strains 186, SD90 or Mock infected by Lamar Fleming and Gabriella Fenkart.
+These data are from running MSD analysis of supernatents from selected vaginal explant samples from seven donors.
+The explants were infected with HSV2 strains 186, SD90 or Mock infected by Lamar Fleming and Gabriella Fenkart and supernatents were collected at 3 timepoints post infection: 3hrs, 8hrs and 24hrs.
 
@@
 
+#Just keep the columns that I care about
+
+explant_all_panels <- explant_all_panels %>%
+  select(Assay,Sample, Sample.Group, Plate.Name, X..Recovery, Concentration, Calc..Concentration, Detection.Range, Fit.Statistic..RSquared,Spot)
 ```
@@
          geom_point(aes(color = Detection.Range),size = 2)+
-  theme(axis.text.x = element_text(angle=25))+
+  theme(axis.text.x = element_text(angle=25, size = 8))+
   labs(y = "Number of Samples", x = "Analyte")+
@@
 
-   geom_point(aes(color = Donor ),size = 1.5, alpha = 0.5)+
+   geom_point(aes(color = Donor ), alpha = 0.5, size = 3)+
   geom_line(aes(group = Donor, color = Donor), alpha = 0.5)+

### 2016-08-23 MSD_panels_1_2_3_HVE_sup.Rmd
First commit of HVE data analysis

---
title: "MSD panels 1, 2 and 3 HVE Sup"
author: "Claire Levy"
date: "August 22, 2016"
output: github_document
---

## Experiment overview

These data are from running MSD analysis of supernatents from selected ex vivo vaginal epithelial cells from three donors. The explants were infected with HSV2 strains 186, SD90 or Mock infected by Lamar Fleming and Gabriella Fenkart and supernatents were collected at 3 timepoints post-infection: 3hrs, 8hrs and 24hrs.

We used 3 different MSD panels containing a total of 20 analytes analyzed for 24 samples. Based on results from running test samples and from running the explant supernatent samples, we decided to dilute all the supernants 1:2 (plate 1) and 1:5 (plate 2) with MSD's diluent 43 before running the plates. All samples were run in duplicate.

```{r read in data, message=FALSE, warning = FALSE, echo = FALSE}

library(dplyr)
library(ggplot2)
library(pander)
library(reshape2)
library(stringr)

#panel 1
# 128 wells x 9 analytes = 1152 obs
HVE_panel_1 <- read.csv("../raw_data/17Aug16_HVE_sup_panel_1_plate_1_and_2.csv", skip = 1)

#panel 2
#128 wells x 8 analytes = 1024 obs
HVE_panel_2 <- read.csv("../raw_data/17Aug16_HVE_sup_panel_2_plate_1_and_2.csv", skip = 1)

#panel 3
#128 wells x 3 analytes = 384 obs

HVE_panel_3 <- read.csv("../raw_data/18Aug16_HVE_sup_panel_3_plate_1_and_2.csv", skip = 1)

#combine data from all 3 panels

HVE_all_panels <-rbind(HVE_panel_1, HVE_panel_2, HVE_panel_3)

#Fix the greek

HVE_all_panels$Assay <-recode(HVE_all_panels$Assay,
                                      "IFN-α2a" = paste("IFN-\U03B1","2a", sep = ""),
                                      "IL-1α" = "IL-1\U03B1",
                                      "MIP-1α"="MIP-1\U03B1",
                                      "TNF-α" = "TNF-\U03B1",
                                      "MIP-3α" = "MIP-3\U03B1",
                                      "MIP-1β" = "MIP-1\U03B2",
                                      "IL-1β" = "IL-1\U03B2",
                                      "IFN-γ" = "IFNg")

#Just keep the columns that I care about

HVE_all_panels <- HVE_all_panels %>%
  select(Assay,Sample, Sample.Group, Plate.Name, X..Recovery, Concentration, Calc..Concentration, Detection.Range, Fit.Statistic..RSquared,Spot)
```

## Checking Standard Recovery and Curves

Here are plots of the % recovery (extrapolated concentration/ known concentration x 100) for the standards that were either within or above the detection range, with lines at the "good recovery" limits of 80 and 120% recovery.

The R^2^ values for goodness-of-fit for the standard curves were all >0.99

```{r standards, echo=FALSE, message = FALSE, warning = FALSE}

HVE_stds_all_panels<-filter(HVE_all_panels, Sample.Group == "Standards")

#which standards are out of det. range?
standardsOutOfRange<-HVE_stds_all_panels %>%
  filter(Detection.Range != "In Detection Range")%>%
  select(Plate.Name,Sample, Assay, Detection.Range, X..Recovery, Calc..Concentration)

#looking at which samples have STDs below the fit curve or below the detection range (i.e. you can't get numbers for % recovery)

HVE_stds_all_panels_below<- HVE_stds_all_panels%>%
  filter(Detection.Range == "Below Fit Curve Range" | Detection.Range == "Below Detection Range") %>%
  group_by(Sample)%>%
  summarise("Number of Samples Below Fit Curve Range/Detection Range" = n())

#The samples that actually give me #'s for percent recovery ( i.e. all but the "below fit curve range" and "below detection range" ones)

HVE_stds_all_panels_percent_recov<- HVE_stds_all_panels %>% 
  arrange(Sample,Assay)%>%
  filter(Detection.Range != "Below Fit Curve Range" & Detection.Range!= "Below Detection Range")%>%
  select(Plate.Name,Sample, Assay,Detection.Range, X..Recovery,Fit.Statistic..RSquared, Concentration, Calc..Concentration)

#plotting percent recovery

ggplot(HVE_stds_all_panels_percent_recov,aes( x = Sample, y = X..Recovery)) +
  geom_point(aes(shape = Detection.Range ),
             size = 2, alpha = 0.5)+
  scale_shape_manual(values = c("Above Fit Curve Range" = 1, 
                                "In Detection Range" = 16, 
                                "Below Detection Range" = 0))+
  facet_wrap(~Assay)+
  theme(axis.text.x = element_text(angle=25))+
  scale_y_continuous(limits = c(0, 265))+
  geom_abline(slope = 0, intercept = c(80, 120))+
  labs(y = "Percent Recovery")+
  theme(axis.text.x = element_text(size = rel(0.8)))

#checking the R squared goodness-of-fit for the standard curves
rSquared<-HVE_stds_all_panels_percent_recov%>%
  group_by(Assay)%>%
  summarize(R_Squared = mean(Fit.Statistic..RSquared))
```

## Caveats about Standards

* Standards with values *below the detection range* (<2.5x the standard deviation of the lowest standard) give unreliable concentrations, so I did not include those in the plots.

* When standards are *below the curve fit* (outside the range of the standard curve), the software does not calculate concentrations so you can't get a percent recovery.

* Most of the Standards that fell into either of these categories were from STD-07 or STD-08 except for five that were STD-06:

`r pander(HVE_stds_all_panels_below)`

```{r break up sample name, echo=FALSE, message = FALSE, warning = FALSE}
#Making a df that contains just the non-standards

HVE_all_panels_noStds <- HVE_all_panels %>%
  filter(Sample.Group!="Standards")

# I want to be able to group the data by timepoint and donor and virus so I need to break up components of the the sample name and make new columns with just the Donor, Timepoint, Virus and Sample_Dilution.

#Split samples names on the "-"
donorTimeVirusDil<-colsplit(HVE_all_panels_noStds$Sample,"-", c("Donor","Timepoint","Virus", "Sample_Dilution") )

#bind the split col back to allPanels

HVE_all_panels_noStds<-cbind(donorTimeVirusDil, HVE_all_panels_noStds)

#Make things factors so they plot nicely
HVE_all_panels_noStds$Donor <- as.factor(HVE_all_panels_noStds$Donor)

HVE_all_panels_noStds$Calc..Concentration <-as.numeric(HVE_all_panels_noStds$Calc..Concentration)

HVE_all_panels_noStds$Virus <- as.factor(HVE_all_panels_noStds$Virus)

HVE_all_panels_noStds$Virus<-factor(HVE_all_panels_noStds$Virus, levels = c("M", "V1","V2"))

HVE_all_panels_noStds$Timepoint<-as.factor(HVE_all_panels_noStds$Timepoint)

HVE_all_panels_noStds$Sample_Dilution<-as.factor(HVE_all_panels_noStds$Sample_Dilution)
```

## HVE Supernatents: Detection Range

The following plots show that the 1:2 dilution of the samples resulted in more samples being in "In Detection Range." Going forward I'll only show data from the 1:2 dilutions.

```{r checking detection range, echo=FALSE, message = FALSE, warning = FALSE}

#What number of samples are in each category of detection? Also adding a column to differentiate which analyte is in which panel, mostly for facetting reasons.

HVE_all_panels_noStds_Below <- HVE_all_panels_noStds%>%
  mutate(Panel = ifelse(str_detect(Plate.Name, "panel_1"),"Panel 1",ifelse(str_detect(Plate.Name,"panel_2"),"Panel 2","Panel 3")))%>%
 group_by(Plate.Name,Sample_Dilution,Panel,Assay, Detection.Range)%>%
  summarise(Number_of_Samples = n())

#Let's plot that

ggplot(HVE_all_panels_noStds_Below, aes(x = Assay, y=Number_of_Samples))+
         geom_point(aes(color = Detection.Range),size = 2)+
  theme(axis.text.x = element_text(angle=25, size = 8))+
  labs(y = "Number of Samples", x = "Analyte")+
  facet_wrap(Sample_Dilution~Panel, scales = "free")+
  ggtitle("Number of Samples in each Detection Range per Analyte")
```

```{r avg reps, echo=FALSE, message = FALSE, warning = FALSE}

### NOTE ### I'm going to filter for just the 1:2 dilution samples here since they gave more in-detection.

HVE_all_panels_noStds_1to2<-HVE_all_panels_noStds %>%
  filter(Sample_Dilution =="1:2")

#I am excluding samples that are "Below Detection Range" because this means that the signal was "statistically indistinguishable from the background or top of the curve"" and ..."The concentrations calculated are inherently unreliable." If one of two reps fell into either of these categories, the remaining replicate will serve as the "average" for that sample.

HVE_all_panels_noStds_1to2_noBelow <- HVE_all_panels_noStds_1to2 %>%
  filter(Detection.Range != "Below Fit Curve Range" & Detection.Range != "Below Detection Range")

#averaging the replicates
avgConc_HVE_all_panels_noStds_1to2_noBelow <- HVE_all_panels_noStds_1to2_noBelow %>%
group_by(Sample, Assay, Donor, Timepoint,Virus)%>%
summarize(avgConc = mean(Calc..Concentration))%>%
  ungroup()

#Looking at the number of samples in each sample "set"(Same Donor and Timepoint). Some are missing V1 or V2. or both!! I don't want the ones with just one sample in the set.

numberInSets<-avgConc_HVE_all_panels_noStds_1to2_noBelow %>%
         group_by(Donor,Timepoint,Assay)%>%
  select(Sample,Donor, Timepoint,Assay)%>%
  summarize(numInSet = n())

#now I'll merge the avgConc df with this numberInSets info created above so I can then filter out samples that have <2 samples in the set

avgConc_HVE_all_panels_noStds_1to2_noBelow <- merge(avgConc_HVE_all_panels_noStds_1to2_noBelow, numberInSets, by = c("Donor", "Timepoint", "Assay"))

#now exclude the samples where there not a "full set", i.e. 1 sample only in the set

avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet <- avgConc_HVE_all_panels_noStds_1to2_noBelow %>%
  filter(numInSet != 1)

```

## Plots of analyte concentrations

Black points show the means.

## Caveats
Samples omitted from the plots include:

* Concentrations below the fit curve range. Values cannot be extrapolated when the data was *below* the fit curve range. However,the MSD software can extrapolate values that are *above* the fit curve range if the curve is linear at the top.

* Samples for which there was only data for a Mock sample and neither of the virus conditions.

Also note:
Some of the samples had one rep that was below detection and another that was in range. For these samples I used the remaining "in range" rep as the "average" for that sample.

```{r  samples plots, echo=FALSE, message = FALSE, warning = FALSE}

#make a list of the analytes

#I'm using a unique vector here instead of the levels of Assay because I don't want to include IFNa2a OR Il-7. There weren't any data points in the detection range for those analytes (so I don't want to try to make a plot of it)

analyteList<-as.list(unique(avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet$Assay))

#write a function to make a plot for each analyte

plotAnalytes <- function (Analyte){
   avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet %>%
     filter(Assay == Analyte)%>%

     ggplot( aes(x = Timepoint, y = avgConc))+

   geom_point(aes(color = Donor ), alpha = 0.5, size = 3)+
  geom_line(aes(group = Donor, color = Donor), alpha = 0.5)+
    stat_summary(fun.y = mean, geom = "point")+
    stat_summary(fun.y = mean, geom = "line",aes(group = 1))+

   scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Detection Range" = 1, "In Detection Range" = 16))+

   scale_y_log10()+

   labs( y = "Concentration")+

  facet_wrap(~Virus, scales = "free_x")+

   theme(axis.text.x = element_text(angle = 25, size = 9))+
     ggtitle(Analyte)
}

lapply(analyteList, FUN = plotAnalytes)

```

### 2016-08-23 MSD_panels_1_2_3_explant_sup_analysis.Rmd
Multiplied the concentrations of the sup by 5 to estimate stock concentration

@@
          geom_point(aes(color = Detection.Range),size = 2)+
-  theme(axis.text.x = element_text(angle=25, size = 8))+
+  theme(axis.text.x = element_text(angle=25, size = 7))+
   labs(y = "Number of Samples", x = "Analyte")+
@@
 
-Black points show the means.
+We did a 1:5 dilution of the samples so I multiplied the concentrations by 5 to get an estimate of the stock suparnatent concentration. Black points show the means.
 
@@
 
+#I'm going to make a column for estimated stock concentration by multiplying the avgConc by 5 (since I did a 1:5 dilution on the samples) so I can better compare with the HVE sup concentrations.
+
+avgConc_explant_all_panels_noStds_noBelow_fullSet <- avgConc_explant_all_panels_noStds_noBelow_fullSet %>%
+  mutate (Estimated_Stock_Conc = avgConc * 5)
+
 #write a function to make a plot for each analyte
@@
 
-     ggplot( aes(x = Timepoint, y = avgConc))+
+     ggplot( aes(x = Timepoint, y = Estimated_Stock_Conc))+
 
@@
 
-   labs( y = "Concentration")+
+   labs( y = "Concentration pg/ml")+
 
@@
 
-   theme(axis.text.x = element_text(angle = 25, size = 9))+
+   theme(axis.text.x = element_text(size = 11))+
      ggtitle(Analyte)

### 2016-08-23 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
multiplied the concentrations by 2 to get an estimate of stock concentration of the sups

--- 2016-08-23 MSD_panels_1_2_3_HVE_sup.Rmd
@@
 ## Plots of analyte concentrations
-
-Black points show the means.
+These samples were diluted 1:2 so I multiplied the concentrations by 2 to estimate the stock concentration of the supernatents. Black points show the means.
 
@@
 
+#I'm going to make a column for estimated stock concentration by multiplying the avgConc by 2(since I did a 1:2 dilution on the samples) so I can better compare with the HVE sup concentrations.
+
+avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet <- avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet %>%
+  mutate (Estimated_Stock_Conc = avgConc * 2)
+
 #write a function to make a plot for each analyte

### 2016-08-23 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
Changed the layout of the individual plots

--- 2016-08-23 MSD_panels_1_2_3_HVE_sup.Rmd
@@
 library(stringr)
+library(swarm)
 
@@
   select(Assay,Sample, Sample.Group, Plate.Name, X..Recovery, Concentration, Calc..Concentration, Detection.Range, Fit.Statistic..RSquared,Spot)
+
 ```
@@
 
-The following plots show that the 1:2 dilution of the samples resulted in more samples being in "In Detection Range." Going forward I'll only show data from the 1:2 dilutions.
+The following plots show that the 1:2 dilution of the samples resulted in more samples being in "In Detection Range."
 
@@
 
-HVE_all_panels_noStds_Below <- HVE_all_panels_noStds%>%
+HVE_all_panels_noStds_NumInEach <- HVE_all_panels_noStds%>%
   mutate(Panel = ifelse(str_detect(Plate.Name, "panel_1"),"Panel 1",ifelse(str_detect(Plate.Name,"panel_2"),"Panel 2","Panel 3")))%>%
@@
 
-ggplot(HVE_all_panels_noStds_Below, aes(x = Assay, y=Number_of_Samples))+
+ggplot(HVE_all_panels_noStds_NumInEach, aes(x = Assay, y=Number_of_Samples))+
          geom_point(aes(color = Detection.Range),size = 2)+
@@
 
-```{r avg reps, echo=FALSE, message = FALSE, warning = FALSE}
+```{r multiplying by dilution factor, echo=FALSE, message = FALSE, warning = FALSE}
 
-### NOTE ### I'm going to filter for just the 1:2 dilution samples here since they gave more in-detection.
+#I'm going to make a column for estimated stock concentration by multiplying the avgConc by 2 or 5 (since I did a 1:2 and 1:5 dilution on the samples) so I can better compare with the explant sup concentrations.
 
-HVE_all_panels_noStds_1to2<-HVE_all_panels_noStds %>%
-  filter(Sample_Dilution =="1:2")
+HVE_all_panels_noStds <- HVE_all_panels_noStds%>%
+  mutate(Estimated_Stock_Conc = ifelse(
+    Sample_Dilution == "1:2", Calc..Concentration * 2,
+    ifelse(Sample_Dilution == "1:5", Calc..Concentration *5, NA)
+  ))
+
+```
+
+```{r exclude belows, echo=FALSE, message = FALSE, warning = FALSE}
 
@@
 
-HVE_all_panels_noStds_1to2_noBelow <- HVE_all_panels_noStds_1to2 %>%
+HVE_all_panels_noStds_noBelow <- HVE_all_panels_noStds %>%
   filter(Detection.Range != "Below Fit Curve Range" & Detection.Range != "Below Detection Range")
+```
+
+## All Samples Estimated Stock Concentrations
+
+* I multiplied by the dilution factors so this is an estimated concentration of the stock samples.
+
+* Note that this is on a log(10) scale; most concentrations are a lot lower than the high ones.
+
+```{r  all sample concentrations, echo=FALSE, message = FALSE, warning = FALSE}
+
+#Plot of all samples and their concentrations
+
+ggplot(HVE_all_panels_noStds_noBelow,aes( x = Timepoint, y = Estimated_Stock_Conc))+
+         geom_point(aes(color = Assay),position = position_swarm (0.5),size = 2)+
+  scale_y_log10()+
+  labs(y = "Estimated Stock Concentration pg/ml")+
+  facet_wrap(~Virus)
+
+```
+
+```{r filter for 1:2, echo=FALSE, message = FALSE, warning = FALSE}
+### NOTE ### I'm going to filter for just the 1:2 dilution samples here since they gave more in-detection.
+
+HVE_all_panels_noStds_1to2_noBelow<-HVE_all_panels_noStds_noBelow %>%
+  filter(Sample_Dilution =="1:2")
+
+```
+
+```{r avg reps, echo=FALSE, message = FALSE, warning = FALSE}
 
@@
 group_by(Sample, Assay, Donor, Timepoint,Virus)%>%
-summarize(avgConc = mean(Calc..Concentration))%>%
+summarize(avgEstConc = mean(Estimated_Stock_Conc))%>%
   ungroup()
@@
 ## Plots of analyte concentrations
-
-Black points show the means.
+Since the 1:2 diluted samples were more likely to be in the detection range, I am only showing data for those.
 
@@
 Also note:
-Some of the samples had one rep that was below detection and another that was in range. For these samples I used the remaining "in range" rep as the "average" for that sample.
+If a sample had one rep that was below detection and another that was in range, I used the remaining "in range" rep as the "average" for that sample.
 
-```{r  samples plots, echo=FALSE, message = FALSE, warning = FALSE}
+```{r  individual analyte plots , echo=FALSE, message = FALSE, warning = FALSE}
+
+ggplot(avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet, aes(x = Timepoint, y = avgEstConc))+
+   geom_point(aes(color = Donor), alpha = 0.75, size = 2)+
+  scale_y_log10()+
+   labs( y = "Estimated Stock Concentration pg/ml")+
+   facet_grid(Assay~Virus, scales = "free_x")+
+   theme(axis.text.x = element_text(size = 11))
+
+################## another way of making the plots #########
 
@@
 
-analyteList<-as.list(unique(avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet$Assay))
+# analyteList<-as.list(unique(avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet$Assay))
 
-#write a function to make a plot for each analyte
-
-plotAnalytes <- function (Analyte){
-   avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet %>%
-     filter(Assay == Analyte)%>%
-
-     ggplot( aes(x = Timepoint, y = avgConc))+
-
-   geom_point(aes(color = Donor ), alpha = 0.5, size = 3)+
-  geom_line(aes(group = Donor, color = Donor), alpha = 0.5)+
-    stat_summary(fun.y = mean, geom = "point")+
-    stat_summary(fun.y = mean, geom = "line",aes(group = 1))+
-
-   scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Detection Range" = 1, "In Detection Range" = 16))+
-
-   scale_y_log10()+
-
-   labs( y = "Concentration")+
-
-  facet_wrap(~Virus, scales = "free_x")+
-
-   theme(axis.text.x = element_text(angle = 25, size = 9))+
-     ggtitle(Analyte)
-}
-
-lapply(analyteList, FUN = plotAnalytes)
+# #write a function to make a plot for each analyte
+# 
+# plotAnalytes <- function (Analyte){
+#    avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet %>%
+#      filter(Assay == Analyte)%>%
+#      
+#      ggplot( aes(x = Timepoint, y = avgEstConc))+
+#    geom_point(aes(color = Donor ), alpha = 0.5, size = 3)+
+#   geom_line(aes(group = Donor, color = Donor), alpha = 0.5)+
+#     stat_summary(fun.y = mean, geom = "point")+
+#     stat_summary(fun.y = mean, geom = "line",aes(group = 1))+
+#    scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Detection Range" = 1, "In Detection Range" = 16))+
+#    scale_y_log10()+
+#    labs( y = "Concentration pg/ml")+
+#   facet_wrap(~Virus, scales = "free_x")+
+#    theme(axis.text.x = element_text(size = 11))+
+#      ggtitle(Analyte)
+# }
+# 
+# 
+# 
+# lapply(analyteList, FUN = plotAnalytes)
 

### 2016-08-23 MSD_panels_1_2_3_explant_sup_analysis.Rmd
Changed the layout of the individual plots

--- 2016-08-23 MSD_panels_1_2_3_explant_sup_analysis.Rmd
@@
 #looking at which samples have STDs below the fit curve or below the detection range (i.e. you can't get numbers for % recovery)
+
 explant_stds_all_panels_below<- explant_stds_all_panels%>%
@@
 #Split samples names on the "-"
-donorTimeVirus<-colsplit(explant_all_panels_noStds$Sample,"-", c("Donor","Timepoint","Virus") )
+donorTimeVirus<-colsplit(explant_all_panels_noStds$Sample,"-", c("Donor","Timepoint","Virus"))
 
-#bind the split col back to allPanels
+#Add a column for Sample dilution (all samples were diluted 1:2)
 
-explant_all_panels_noStds<-cbind(donorTimeVirus, explant_all_panels_noStds)
+donorTimeVirusDil <- donorTimeVirus %>%
+  mutate(Sample_Dilution = "1:2")
+
+#bind the donorTimeVirusDil to allPanels
+
+explant_all_panels_noStds<-cbind(donorTimeVirusDil, explant_all_panels_noStds)
 
@@
 #What number of samples are in each category of detection?
-explant_all_panels_noStds_Below <- explant_all_panels_noStds%>%
+explant_all_panels_noStds_NumInEach <- explant_all_panels_noStds%>%
   mutate(Panel = ifelse(str_detect(Plate.Name, "panel_1"),"Panel 1",ifelse(str_detect(Plate.Name,"panel_2"),"Panel 2","Panel 3")))%>%
@@
 
-ggplot(explant_all_panels_noStds_Below, aes(x = Assay, y=Number_of_Samples))+
+ggplot(explant_all_panels_noStds_NumInEach, aes(x = Assay, y=Number_of_Samples))+
          geom_point(aes(color = Detection.Range),size = 2)+
-  theme(axis.text.x = element_text(angle=25, size = 8))+
+  theme(axis.text.x = element_text(angle=25, size = 7))+
   labs(y = "Number of Samples", x = "Analyte")+
@@
 
-```{r avg reps, echo=FALSE, message = FALSE, warning = FALSE}
+```{r multiplying by dilution factor, echo=FALSE, message = FALSE, warning = FALSE}
 
-#some of the samples had one rep that was below detection, so I am excluding those so I will be able to use the remaining rep as the "average" for that sample.
+#I'm going to make a column for estimated stock concentration by multiplying the avgConc by 2 (since I did a 1:2 dilution on the samples) so I can better compare with the HVE sup concentrations.
+
+explant_all_panels_noStds <- explant_all_panels_noStds%>%
+  mutate(Estimated_Stock_Conc =  Calc..Concentration * 2)
+
+```
+
+```{r exclude belows, echo=FALSE, message = FALSE, warning = FALSE}
 
@@
   filter(Detection.Range != "Below Fit Curve Range" & Detection.Range != "Below Detection Range")
+```
+
+## All Samples Estimated Stock Concentrations
+
+* I multiplied by the dilution factor so this is an estimated concentration of the stock samples.
+
+* Note that this is on a log(10) scale; most concentrations are a lot lower than the high ones.
+
+```{r  all sample concentrations, echo=FALSE, message = FALSE, warning = FALSE}
+
+#Plot of all samples and their concentrations
+
+ggplot(explant_all_panels_noStds_noBelow,aes( x = Timepoint, y = Estimated_Stock_Conc))+
+         geom_point(aes(color = Assay),position = position_swarm (0.5),size = 2)+
+  scale_y_log10()+
+  labs(y = "Estimated Stock Concentration pg/ml")+
+  facet_wrap(~Virus)
+
+```
+
+```{r averaging reps, echo=FALSE, message = FALSE, warning = FALSE}
 
 #averaging the replicates
+
+#some of the samples had one rep that was below detection, so I am excluding those so I will be able to use the remaining rep as the "average" for that sample.
+
 avgConc_explant_all_panels_noStds_noBelow <- explant_all_panels_noStds_noBelow %>%
 group_by(Sample, Assay, Donor, Timepoint,Virus)%>%
-summarize(avgConc = mean(Calc..Concentration))%>%
+summarize(avgEstConc = mean(Estimated_Stock_Conc))%>%
   ungroup()
@@
 Also note:
-Some of the samples had one rep that was below detection and another that was in range. For these samples I used the remaining "in range" rep as the "average" for that sample.
+If a sample had one rep that was below detection and another that was in range, I used the remaining "in range" rep as the "average" for that sample.
 
-```{r  samples plots, echo=FALSE, message = FALSE, warning = FALSE}
+```{r  individual analyte plots, echo=FALSE, message = FALSE, warning = FALSE}
 
-#make a list of the analytes
+ggplot(avgConc_explant_all_panels_noStds_noBelow_fullSet, aes(x = Timepoint, y = avgEstConc))+
+   geom_point(aes(color = Donor),position = position_swarm(0.07), alpha = 0.75, size = 1.5)+
+  scale_y_log10()+
+   labs( y = "Estimated Stock Concentration pg/ml")+
+   facet_grid(Assay~Virus, scales = "free_x")+
+   theme(axis.text.x = element_text(size = 11))
 
-#I'm using a unique vector here instead of the levels of Assay because I don't want to include IFNa2a OR Il-7.
+################## another way of making the plots #####
+
+# #make a list of the analytes
+# 
+# #I'm using a unique vector here instead of the levels of Assay because I don't want to include IFNa2a OR Il-7.
 There weren't any data points in the detection range for those analytes (so I don't want to try to make a plot of it)
-
-analyteList<-as.list(unique(avgConc_explant_all_panels_noStds_noBelow$Assay))
-
-#write a function to make a plot for each analyte
-
-plotAnalytes <- function (Analyte){
-   avgConc_explant_all_panels_noStds_noBelow_fullSet %>%
-     filter(Assay == Analyte)%>%
-
-     ggplot( aes(x = Timepoint, y = avgConc))+
-
-   geom_point(aes(color = Donor ), alpha = 0.5, size = 3)+
-  geom_line(aes(group = Donor, color = Donor), alpha = 0.5)+
-    stat_summary(fun.y = mean, geom = "point")+
-    stat_summary(fun.y = mean, geom = "line",aes(group = 1))+
-
-   scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Detection Range" = 1, "In Detection Range" = 16))+
-
-   scale_y_log10()+
-
-   labs( y = "Concentration")+
-
-  facet_wrap(~Virus, scales = "free_x")+
-
-   theme(axis.text.x = element_text(angle = 25, size = 9))+
-     ggtitle(Analyte)
-}
-
-lapply(analyteList, FUN = plotAnalytes)
+# 
+# 
+# analyteList<-as.list(unique(avgConc_explant_all_panels_noStds_noBelow$Assay))
+# 
+# 
+# 
+# 
+# #write a function to make a plot for each analyte
+# 
+# plotAnalytes <- function (Analyte){
+#    avgConc_explant_all_panels_noStds_noBelow_fullSet %>%
+#      filter(Assay == Analyte)%>%
+#      
+#      ggplot( aes(x = Timepoint, y = Estimated_Stock_Conc))+
+#    
+#    geom_point(aes(color = Donor ), alpha = 0.5, size = 3)+
+#   geom_line(aes(group = Donor, color = Donor), alpha = 0.5)+
+#     stat_summary(fun.y = mean, geom = "point")+
+#     stat_summary(fun.y = mean, geom = "line",aes(group = 1))+
+#    
+#    scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Detection Range" = 1, "In Detection Range" = 16))+
+#   
+#    scale_y_log10()+
+#    
+#    labs( y = "Concentration pg/ml")+
+#    
+#   facet_wrap(~Virus, scales = "free_x")+
+#    
+#    theme(axis.text.x = element_text(size = 11))+
+#      ggtitle(Analyte)
+# }
+# 
+# 
+# 
+# lapply(analyteList, FUN = plotAnalytes)
 

### 2016-08-23 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
adjusted plot sizes for HTML

@@
                                       "IL-1β" = "IL-1\U03B2",
-                                      "IFN-γ" = "IFNg")
+                                      "IFN-γ" = "IFNg",
+                              "MIG/CXCL9" = "MIG")
 
@@
 
-```{r checking detection range, echo=FALSE, message = FALSE, warning = FALSE}
+```{r checking detection range,fig.height = 6, fig.width = 10, echo=FALSE, message = FALSE, warning = FALSE}
 
@@
 
-```{r  individual analyte plots , echo=FALSE, message = FALSE, warning = FALSE}
+```{r  individual analyte plots,fig.width = 8, fig.height= 13 , echo=FALSE, message = FALSE, warning = FALSE}
 
 ggplot(avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet, aes(x = Timepoint, y = avgEstConc))+
-   geom_point(aes(color = Donor), alpha = 0.75, size = 2)+
+   geom_point(aes(color = Donor), alpha = 0.75, size = 3)+
   scale_y_log10()+

### 2016-08-23 MSD_panels_1_2_3_explant_sup_analysis.Rmd
adjusted plot sizes for HTML

@@
                                       "IL-1β" = "IL-1\U03B2",
-                                      "IFN-γ" = "IFNg")
+                                      "IFN-γ" = "IFNg",
+                                  "MIG/CXCL9" = "MIG")
 
@@
 
-```{r checking detection range, echo=FALSE, message = FALSE, warning = FALSE}
+```{r checking detection range,fig.height = 5, fig.width = 10, echo=FALSE, message = FALSE, warning = FALSE}
 
@@
          geom_point(aes(color = Detection.Range),size = 2)+
-  theme(axis.text.x = element_text(angle=25, size = 7))+
+  theme(axis.text.x = element_text(angle=25, size = 8))+
   labs(y = "Number of Samples", x = "Analyte")+
@@
 
-We did a 1:5 dilution of the samples so I multiplied the concentrations by 5 to get an estimate of the stock suparnatent concentration. Black points show the means.
+We did a 1:5 dilution of the samples so I multiplied the concentrations by 5 to get an estimate of the stock suparnatent concentration.
 
@@
 
-```{r  individual analyte plots, echo=FALSE, message = FALSE, warning = FALSE}
+```{r  individual analyte plots, fig.width = 8, fig.height= 13, echo=FALSE}
 
 ggplot(avgConc_explant_all_panels_noStds_noBelow_fullSet, aes(x = Timepoint, y = avgEstConc))+
-   geom_point(aes(color = Donor),position = position_swarm(0.07), alpha = 0.75, size = 1.5)+
+   geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
   scale_y_log10()+

### 2016-09-20 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
Changed the below det values to 0.01, subtracted mock from the virus data

@@
 ---
-title: "MSD panels 1, 2 and 3 HVE Sup"
-author: "Claire Levy"
-date: "August 22, 2016"
-output: github_document
+title: "MSD panels 1, 2 and 3 HVE Supernatant"
+output: pdf_document
 ---
@@
 
-```{r standards, echo=FALSE, message = FALSE, warning = FALSE}
+```{r standards,fig.height = 7, fig.width = 12, echo=FALSE, message = FALSE, warning = FALSE}
 
@@
   labs(y = "Percent Recovery")+
-  theme(axis.text.x = element_text(size = rel(0.8)))
+  theme(axis.text.x = element_text(size = 9))+
+  ggtitle("Percent Recovery of Standards")
 
@@
 
-```{r checking detection range,fig.height = 6, fig.width = 10, echo=FALSE, message = FALSE, warning = FALSE}
+```{r checking detection range, fig.height=8, fig.width =12, echo=FALSE, message = FALSE, warning = FALSE}
 
@@
          geom_point(aes(color = Detection.Range),size = 2)+
-  theme(axis.text.x = element_text(angle=25, size = 8))+
+  theme(axis.text.x = element_text(angle=25, size = 11))+
   labs(y = "Number of Samples", x = "Analyte")+
@@
 
-## All Samples Estimated Stock Concentrations
+## Estimated Sample Concentration Distribution
 
@@
 
-```{r  all sample concentrations, echo=FALSE, message = FALSE, warning = FALSE}
+```{r  all sample concentrations,fig.height = 6, fig.with = 10,  echo=FALSE, message = FALSE, warning = FALSE}
 
@@
   labs(y = "Estimated Stock Concentration pg/ml")+
-  facet_wrap(~Virus)
+  facet_wrap(~Virus)+
+  ggtitle("Estimated Sample Concentration Distribution by Treatment and Timepoint")
 
@@
 
-## Plots of analyte concentrations
+## Plots of Analyte Concentrations by Donor, Timepoint and Condition
+
 Since the 1:2 diluted samples were more likely to be in the detection range, I am only showing data for those.
@@
 
-```{r  individual analyte plots,fig.width = 8, fig.height= 13 , echo=FALSE, message = FALSE, warning = FALSE}
+```{r  individual analyte plots, echo=FALSE,fig.width=8, fig.height = 10, message = FALSE, warning = FALSE}
 
@@
    facet_grid(Assay~Virus, scales = "free_x")+
-   theme(axis.text.x = element_text(size = 11))
+   theme(axis.text.x = element_text(size = 11))+
+  ggtitle("Analyte Concentrations by Donor, Timepoint and Condition")
 

### 2016-09-20 MSD_panels_1_2_3_explant_sup_analysis.Rmd
Changed the below det values to 0.01, subtracted mock from the virus data

@@
 ---
-title: "MSD panels 1, 2 and 3 Explant Sup"
-author: "Claire Levy"
-date: "August 12, 2016"
-output: github_document
+title: "MSD panels 1, 2 and 3 Explant Supernatant"
+output: pdf_document
 ---
@@
 #remember that this data table contains data from the 17Aug16_HVE_sup_panel_1_plate_1.
-I am only going to use the Standards data from that plate and not the standards from either of the plates for panel 1 run 10Aug16 since we had what appears to be a pipetting error (2nd replicate is more concentration for all wells).
+I am only going to use the Standards data from that plate (not the data) and not the standards from either of the plates for panel 1 run 10Aug16 since we had what appears to be a pipetting error (2nd replicate is more concentrated for all wells).
 
@@
 
-#Add a column for Sample dilution (all samples were diluted 1:2)
+#Add a column for Sample dilution (all samples were diluted 1:5)
 
 donorTimeVirusDil <- donorTimeVirus %>%
-  mutate(Sample_Dilution = "1:2")
+  mutate(Sample_Dilution = "1:5")
 
@@
 
-#checking to make sure that only the samples that SHOULD have 1 rep, show up as having just 1 rep (sample 324-T2-V1 where I ran out of sample before I could do the 2nd rep)
+#checking to make sure that only the samples that SHOULD have 1 rep, show up as having just 1 rep (sample 324-T2-V1 where I ran out of sample before I could do the 2nd rep for panel 1.
+there wasnt enough of this sample to use any for panel 2 :( )
 
@@
 
-```{r checking detection range,fig.height = 5, fig.width = 10, echo=FALSE, message = FALSE, warning = FALSE}
+```{r checking detection range,fig.height = 4, fig.width = 12, echo=FALSE, message = FALSE, warning = FALSE}
 
@@
 
+#Were there any samples where none of the assays seemed to have worked 
+#i.e. All 40 were below?
+explant_all_panels_noStds_NumInEach_bySample <- explant_all_panels_noStds%>%
+ group_by(Sample, Detection.Range)%>%
+  summarise(Number_of_Reps = n())%>%
+  filter(str_detect(Detection.Range, "Below"))%>%
+  group_by(Sample)%>%
+  summarise(Number_of_Belows = sum(Number_of_Reps))
+#None of these numbers are 40, so it is ok.
+
 ```
@@
 
-#I'm going to make a column for estimated stock concentration by multiplying the avgConc by 2 (since I did a 1:2 dilution on the samples) so I can better compare with the HVE sup concentrations.
+#I'm going to make a column for estimated stock concentration by multiplying the avgConc by 5 (since I did a 1:5 dilution on the samples) so I can better compare with the HVE sup concentrations.
 
 explant_all_panels_noStds <- explant_all_panels_noStds%>%
-  mutate(Estimated_Stock_Conc =  Calc..Concentration * 2)
+  mutate(Estimated_Stock_Conc =  Calc..Concentration * 5)
 
@@
 
-```{r exclude belows, echo=FALSE, message = FALSE, warning = FALSE}
+```{r adjust belows, echo=FALSE, message = FALSE, warning = FALSE}
 
-#I am also excluding samples that are "Below Detection Range" because this means that the signal was "statistically indistinguishable from the background or top of the curve"" and ..."The concentrations calculated are inherently unreliable."
+#Samples that are "Below Detection Range" means that the signal was "statistically indistinguishable from the background or top of the curve"" and ..."The concentrations calculated are inherently unreliable."
 
-explant_all_panels_noStds_noBelow <- explant_all_panels_noStds %>%
-  filter(Detection.Range != "Below Fit Curve Range" & Detection.Range != "Below Detection Range")
+#HOWEVER, we still want to show the fact that those samples had really low concentrations. So, I am going to set all those samples to 0.001 so they show up on plots.
+
+#which ones were those again?
+
+explant_belows <- explant_all_panels_noStds %>%
+  filter(str_detect(Detection.Range, "Below"))%>%
+  group_by(Donor, Timepoint, Assay)%>%
+  summarize(Number_of_samples = n())
+
+#I'm going to set all of the "belows" to 0.001
+
+explant_all_panels_noStds_AdjBelow <- explant_all_panels_noStds %>%
+mutate(Estimated_Stock_Conc = ifelse(str_detect(Detection.Range,"Below"),0.01, Estimated_Stock_Conc))
+
 ```
 
-## All Samples Estimated Stock Concentrations
+##Estimated Sample Concentration Distribution
 
@@
 
-```{r  all sample concentrations, echo=FALSE, message = FALSE, warning = FALSE}
+```{r  all sample concentrations,fig.height = 6, fig.with = 10, echo=FALSE, message = FALSE, warning = FALSE, results = 'asis'}
 
 #Plot of all samples and their concentrations
+#Note that this includes the samples that manually set to 0.01
 
-ggplot(explant_all_panels_noStds_noBelow,aes( x = Timepoint, y = Estimated_Stock_Conc))+
+ggplot(explant_all_panels_noStds_AdjBelow,aes( x = Timepoint, y = Estimated_Stock_Conc))+
          geom_point(aes(color = Assay),position = position_swarm (0.5),size = 2)+
@@
   labs(y = "Estimated Stock Concentration pg/ml")+
-  facet_wrap(~Virus)
+  facet_wrap(~Virus)+
+  ggtitle("Estimated Sample Concentration Distribution by Treatment and Timepoint")
+
+cat("\n\n\\pagebreak\n")
 
@@
 
+#First change the "Below..." entries to just say "Below", since I don't care which "Below..." it was for my purposes.
+
+explant_all_panels_noStds_AdjBelow <- explant_all_panels_noStds_AdjBelow %>%
+  mutate(Detection.Range = ifelse(str_detect(Detection.Range, "Below"), "Below", as.character(Detection.Range)))
+
+#which samples have discordant replicates in terms of det.range?
+repsInRange<-explant_all_panels_noStds_AdjBelow%>%
+  group_by(Sample, Assay, Detection.Range)%>%
+  summarize(Number_Reps = n())
+
+#merge the repsInRange data with the explant_all_panels_noStds_AdjBelow data by the Sample and Assay
+
+#Result is a df where the samples with dischordant replicates in terms of Detection.Range.
+The samples with "1" in the Number_of_Reps colimn are the ones where I want to use the data from the rep that was in/above range for the avg, NOT the average between the in-range rep and the below-range rep.
+Recall that I assigned all below range/curve reps a value of 0.01.
+If both reps were below, ok to use calculate the average of them.
+
+# select fewer cols for readability.
+
+mergeRepsInRange <- explant_all_panels_noStds_AdjBelow %>%
+  merge(.,repsInRange, by = c("Sample", "Assay", "Detection.Range"))%>%
+  select(Sample,Timepoint,Donor,Virus,Assay,Detection.Range, Estimated_Stock_Conc, Number_Reps, Plate.Name)
+
+#HOWEVER: I know that for sample 324-T2-V1, I only had one replicate for the assays on panel 2, so for just this sample, I don't want to eliminate a rep that meets this exclusion criteria (IL-12p70) because I don't actually have a good rep to use as the average.
+I'll add it back in after filtering.
+
+T2V1324<-mergeRepsInRange %>%
+  filter(Sample == "324-T2-V1", Assay =="IL-12p70")
+
+#The  rows of data that I want will fit the following criteria:
+
+#Detection.Range = Below and Number_reps = 2 (both reps are below)
+#Detection.Range != Below and Number_reps = 2 (both reps are good)
+#Detection.Range != Below and Number_reps = 1 (The in/above range rep from a discordant set)
+
+#I am doing this by filtering for the things I want to keep, then combining those dfs, rather than writing one expression to say what to exclude. (mostly because I couldnt figure it out...)
+
+bothBad <- mergeRepsInRange %>%
+  filter(Detection.Range =="Below"& Number_Reps ==2)
+
+discordGood<- mergeRepsInRange %>%
+  filter(Detection.Range != "Below"& Number_Reps ==1)
+
+bothGood <- mergeRepsInRange %>%
+  filter(Detection.Range != "Below"& Number_Reps == 2)
+
+#combine all the reps
+
+OKtoAvg <- rbind(bothBad,discordGood, bothGood,T2V1324)
+
 #averaging the replicates
-
-#some of the samples had one rep that was below detection, so I am excluding those so I will be able to use the remaining rep as the "average" for that sample.
-
-avgConc_explant_all_panels_noStds_noBelow <- explant_all_panels_noStds_noBelow %>%
+avgConc_explant_all_panels_noStds_AdjBelow <- OKtoAvg %>%
 group_by(Sample, Assay, Donor, Timepoint,Virus)%>%
@@
 
-#Looking at the number of samples in each sample "set"(Same Donor and Timepoint). Some are missing V1 or V2. or both!! I don't want the ones with just one sample in the set.
+```
 
-numberInSets<-avgConc_explant_all_panels_noStds_noBelow %>%
+```{r  samples in sets, fig.width = 8, fig.height= 13, echo=FALSE}
+
+#Looking at the number of samples in each sample "set"(Same Donor and Timepoint).
+
+numberInSets<-avgConc_explant_all_panels_noStds_AdjBelow %>%
          group_by(Donor,Timepoint,Assay)%>%
@@
 
-#now I'll merge the avgConc df with this numberInSets info created above so I can then filter out samples that have <2 samples in the set
+#now I'll merge the avgConc df with this numberInSets info so I know which conditions are in each set.
+avgConc_explant_all_panels_noStds_AdjBelow <- merge(avgConc_explant_all_panels_noStds_AdjBelow, numberInSets, by = c("Donor", "Timepoint", "Assay"))
 
-avgConc_explant_all_panels_noStds_noBelow <- merge(avgConc_explant_all_panels_noStds_noBelow, numberInSets, by = c("Donor", "Timepoint", "Assay"))
-
-#now exclude the samples where there not a "full set", i.e. 1 sample only in the set
-
-avgConc_explant_all_panels_noStds_noBelow_fullSet <- avgConc_explant_all_panels_noStds_noBelow %>%
-  filter(numInSet != 1)
+# For 324-T2, there are ONLY mock samples
+# There are no samples with just V1 and/or V2 and no mock
 
@@
 
-## Plots of analyte concentrations
+```{r  subtract mock, fig.width = 8, fig.height= 13, echo=FALSE}
 
+#I want to subtract out the mock data from the corresponding v1 and V2 data.
+
+#extract out Mock samples
+
+mock <- avgConc_explant_all_panels_noStds_AdjBelow %>%
+  filter(Virus == "M")
+
+# remove mock from the rest of the data
+
+no_mock_avgConc_explant_all_panels_noStds_AdjBelow <-
+  avgConc_explant_all_panels_noStds_AdjBelow %>%
+  filter(Virus !="M")
+
+mergeMock<-merge(no_mock_avgConc_explant_all_panels_noStds_AdjBelow, mock, by = c("Donor", "Assay", "Timepoint"), all.x = TRUE, all.y = TRUE)
+
+#check for NA's (places where there was a mock sample but no corresponding V1 or V2)
+whichNAs<-mergeMock[which(is.na(mergeMock$Sample.x)),]
+#This is just donor 324 Il1ra, mig and rantes, which we already know has just mock.
+# Remove the NAs for tidyness
+
+mergeMock<-mergeMock%>%
+  na.omit()%>%
+  select(Donor, Assay, Timepoint, Sample.x, Virus.x, avgEstConc.x, numInSet.x, avgEstConc.y)
+
+#Fix up the colnames
+
+names(mergeMock)[4:8]<- c("Sample","Virus","avgEstConcVirus","numInSet","avgEstConcMock")
+
+#Do the subtraction
+
+mergeMock<- mergeMock %>%
+  mutate(minusMockEstConc = avgEstConcVirus-avgEstConcMock)
+
+#When the experimental and mock samples were both "below", we get zero in the minusMockEstConc column, and that won't show up on a log scale.
+So I'm setting all zeros in the minusMockEstConc column to 0.01
+
+mergeMock <- mergeMock %>%
+  mutate(minusMockEstConc = ifelse(minusMockEstConc <= 0, 0.01, minusMockEstConc))
+
+anyNA(mergeMock)
+```
+
+## Plots of Analyte Concentrations by Donor, Timepoint and Condition
 We did a 1:5 dilution of the samples so I multiplied the concentrations by 5 to get an estimate of the stock suparnatent concentration.
@@
 
-* Concentrations below the fit curve range.
-Values cannot be extrapolated when the data was *below* the fit curve range.
-However,the MSD software can extrapolate values that are *above* the fit curve range if the curve is linear at the top.
-
-* Samples for which there was only data for a Mock sample and neither of the virus conditions.
+* Samples for which there was only data for a Mock sample and neither of the virus conditions (Donor 324, IL-1RA, MIG, RANTES)
 
@@
 
-ggplot(avgConc_explant_all_panels_noStds_noBelow_fullSet, aes(x = Timepoint, y = avgEstConc))+
+ggplot(mergeMock, aes(x = Timepoint, y = minusMockEstConc))+
    geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
@@
    facet_grid(Assay~Virus, scales = "free_x")+
-   theme(axis.text.x = element_text(size = 11))
-
-################## another way of making the plots #####
-
-# #make a list of the analytes
-# 
-# #I'm using a unique vector here instead of the levels of Assay because I don't want to include IFNa2a OR Il-7.
-There weren't any data points in the detection range for those analytes (so I don't want to try to make a plot of it)
-# 
-# 
-# analyteList<-as.list(unique(avgConc_explant_all_panels_noStds_noBelow$Assay))
-# 
-# 
-# 
-# 
-# #write a function to make a plot for each analyte
-# 
-# plotAnalytes <- function (Analyte){
-#    avgConc_explant_all_panels_noStds_noBelow_fullSet %>%
-#      filter(Assay == Analyte)%>%
-#      
-#      ggplot( aes(x = Timepoint, y = Estimated_Stock_Conc))+
-#    
-#    geom_point(aes(color = Donor ), alpha = 0.5, size = 3)+
-#   geom_line(aes(group = Donor, color = Donor), alpha = 0.5)+
-#     stat_summary(fun.y = mean, geom = "point")+
-#     stat_summary(fun.y = mean, geom = "line",aes(group = 1))+
-#    
-#    scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Detection Range" = 1, "In Detection Range" = 16))+
-#   
-#    scale_y_log10()+
-#    
-#    labs( y = "Concentration pg/ml")+
-#    
-#   facet_wrap(~Virus, scales = "free_x")+
-#    
-#    theme(axis.text.x = element_text(size = 11))+
-#      ggtitle(Analyte)
-# }
-# 
-# 
-# 
-# lapply(analyteList, FUN = plotAnalytes)
+   theme(axis.text.x = element_text(size = 11))+
+  ggtitle("Analyte Concentrations by Donor, Timepoint and Condition")
 

### 2016-09-21 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
Changed the way I was subtracting the mock so I could plot on a linear scale, added a plot of just the microarray-interesting subset of donors.

@@
 
-```{r exclude belows, echo=FALSE, message = FALSE, warning = FALSE}
+```{r adjust belows, echo=FALSE, message = FALSE, warning = FALSE}
+#Samples that are "Below Detection Range" means that the signal was "statistically indistinguishable from the background or top of the curve"" and ..."The concentrations calculated are inherently unreliable."
 
-#I am excluding samples that are "Below Detection Range" because this means that the signal was "statistically indistinguishable from the background or top of the curve"" and ..."The concentrations calculated are inherently unreliable." If one of two reps fell into either of these categories, the remaining replicate will serve as the "average" for that sample.
+#HOWEVER, we still want to show the fact that those samples had really low concentrations. So, I am going to set all those samples to 0.001 so they show up on plots.
 
-HVE_all_panels_noStds_noBelow <- HVE_all_panels_noStds %>%
-  filter(Detection.Range != "Below Fit Curve Range" & Detection.Range != "Below Detection Range")
+HVE_all_panels_noStds_AdjBelow <- HVE_all_panels_noStds %>%
+  mutate(Estimated_Stock_Conc = ifelse(str_detect(Detection.Range,"Below"),0.01, Estimated_Stock_Conc))
 ```
@@
 
+```{r filter for 1:2, echo=FALSE, message = FALSE, warning = FALSE}
+### NOTE ### I'm going to filter for just the 1:2 dilution samples here since they gave more in-detection.
+
+HVE_all_panels_noStds_AdjBelow_short<-HVE_all_panels_noStds_AdjBelow %>%
+  filter(Sample_Dilution =="1:2", Detection.Range != "Above Fit Curve Range")
+
+#There were a few samples where the 1:2 dilution was above the fit curve range but it was in range for 1:5.
+I am going to extract those samples and add them to the df I made above, along with their corresponding Mock samples (which are also in range
+
+IL1RA_4_T3<-HVE_all_panels_noStds_AdjBelow%>%
+  filter(Assay == "IL1-RA", Donor == 4, Timepoint == "T3", Sample_Dilution=="1:5")
+
+IL1RA_2_T2<-HVE_all_panels_noStds_AdjBelow%>%
+  filter(Assay == "IL1-RA", Donor == 2, Timepoint == "T2", Sample_Dilution=="1:5")
+
+HVE_all_panels_noStds_AdjBelow_short <- rbind(HVE_all_panels_noStds_AdjBelow_short,IL1RA_4_T3,IL1RA_2_T2)
+
+```
+
 ```{r  all sample concentrations,fig.height = 6, fig.with = 10,  echo=FALSE, message = FALSE, warning = FALSE}
@@
 
-ggplot(HVE_all_panels_noStds_noBelow,aes( x = Timepoint, y = Estimated_Stock_Conc))+
+ggplot(HVE_all_panels_noStds_AdjBelow_short,aes( x = Timepoint, y = Estimated_Stock_Conc))+
          geom_point(aes(color = Assay),position = position_swarm (0.5),size = 2)+
@@
 
-```{r filter for 1:2, echo=FALSE, message = FALSE, warning = FALSE}
-### NOTE ### I'm going to filter for just the 1:2 dilution samples here since they gave more in-detection.
-
-HVE_all_panels_noStds_1to2_noBelow<-HVE_all_panels_noStds_noBelow %>%
-  filter(Sample_Dilution =="1:2")
-
-```
-
 ```{r avg reps, echo=FALSE, message = FALSE, warning = FALSE}
 
-#averaging the replicates
-avgConc_HVE_all_panels_noStds_1to2_noBelow <- HVE_all_panels_noStds_1to2_noBelow %>%
-group_by(Sample, Assay, Donor, Timepoint,Virus)%>%
+#First change the "Below..." entries to just say "Below", since I don't care which "Below..." it was for my purposes.
+
+HVE_all_panels_noStds_AdjBelow_short <- HVE_all_panels_noStds_AdjBelow_short %>%
+  mutate(Detection.Range = ifelse(str_detect(Detection.Range, "Below"), "Below", as.character(Detection.Range)))
+
+#which samples have discordant replicates in terms of det.range?
+repsInRange<-HVE_all_panels_noStds_AdjBelow_short%>%
+  group_by(Sample, Assay, Detection.Range)%>%
+  summarize(Number_Reps = n())
+
+#merge the repsInRange data with the HVE_all_panels_noStds_1to2_AdjBelow data by the Sample and Assay
+
+#Result is a df where the samples with dischordant replicates in terms of Detection.Range.
+The samples with "1" in the Number_of_Reps colimn are the ones where I want to use the data from the rep that was in/above range for the avg, NOT the average between the in-range rep and the below-range rep.
+Recall that I assigned all below range/curve reps a value of 0.01.
+If both reps were below, ok to use calculate the average of them.
+
+# select fewer cols for readability.
+
+mergeRepsInRange <- HVE_all_panels_noStds_AdjBelow_short %>%
+  merge(.,repsInRange, by = c("Sample", "Assay", "Detection.Range"))%>%
+  select(Sample,Timepoint,Donor,Virus,Assay,Detection.Range, Estimated_Stock_Conc, Number_Reps, Plate.Name, Sample_Dilution)
+
+#The  rows of data that I want will fit the following criteria:
+
+#Detection.Range = Below and Number_reps = 2 (both reps are below)
+#Detection.Range != Below and Number_reps = 2 (both reps are good)
+#Detection.Range != Below and Number_reps = 1 (The in/above range rep from a discordant set)
+
+#I am doing this by filtering for the things I want to keep, then combining those dfs, rather than writing one expression to say what to exclude. (mostly because I couldnt figure it out...)
+
+bothBad <- mergeRepsInRange %>%
+  filter(Detection.Range =="Below"& Number_Reps ==2)
+
+discordGood<- mergeRepsInRange %>%
+  filter(Detection.Range == "In Detection Range" & Number_Reps ==1)
+
+bothGood <- mergeRepsInRange %>%
+  filter(Detection.Range == "In Detection Range"& Number_Reps == 2)
+
+#combine all the reps, including the data the I extracted from the 1:5 dilution for IL1-RA
+
+OKtoAvg <- rbind(bothBad,discordGood, bothGood)
+
+#averaging the replicates. Should have 944 reps - 26 single reps = 918 reps
+
+#918 reps /2 =459 averages + 26 averages from single reps = 485 values
+avgConc <- OKtoAvg %>%
+group_by(Sample, Assay, Donor, Timepoint,Virus, Sample_Dilution)%>%
 summarize(avgEstConc = mean(Estimated_Stock_Conc))%>%
@@
 
-numberInSets<-avgConc_HVE_all_panels_noStds_1to2_noBelow %>%
+numberInSets<-avgConc %>%
          group_by(Donor,Timepoint,Assay)%>%
@@
 
-#now I'll merge the avgConc df with this numberInSets info created above so I can then filter out samples that have <2 samples in the set
+#all complete sets!
 
-avgConc_HVE_all_panels_noStds_1to2_noBelow <- merge(avgConc_HVE_all_panels_noStds_1to2_noBelow, numberInSets, by = c("Donor", "Timepoint", "Assay"))
+```
 
-#now exclude the samples where there not a "full set", i.e. 1 sample only in the set
+```{r  subtract mock, fig.width = 8, fig.height= 13, echo=FALSE}
 
-avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet <- avgConc_HVE_all_panels_noStds_1to2_noBelow %>%
-  filter(numInSet != 1)
+#I want to subtract out the mock data from the corresponding v1 and V2 data.
+
+#extract out Mock samples
+
+mock <- avgConc %>%
+  filter(Virus == "M")
+
+# remove mock from the rest of the data
+
+no_mock_avgConc <-
+  avgConc %>%
+  filter(Virus !="M")
+
+mergeMock<-merge(no_mock_avgConc, mock, by = c("Donor","Timepoint","Assay", "Sample_Dilution"), all.x = TRUE, all.y = TRUE)
+
+#check for NA's (places where there was a mock sample but no corresponding V1 or V2)
+whichNAs<-mergeMock[which(is.na(mergeMock$Sample.x)),]
+
+#no NA's!
+
+#Fix up the colnames:
+
+names(mergeMock)[c(5:7,10)]<- c("Sample","Virus","avgEstConcVirus","avgEstConcMock")
+
+#When the experimental and mock samples were both "below", we get zero in the minusMockEstConc column, and that won't show up on a log scale.
+So I'm setting all zeros in the minusMockEstConc column to 0.01.
+
+#Also, in cases where the Mock values are > Virus values, we would get a negative number for the result and not be able to plot it on a log scale.
+So I will take the log of each, then subtract, then plot on a linear scale, so I can have the negatives there.
+
+#Recall that log(a/b) = log(a)-log(b)
+#so if Mock = 100 and V = 10, log 100/10 = log (0.1) = -1
+# and log(100)-log(10) = 1-2 = -1
+
+mergeMock <- mergeMock %>%
+  mutate(logMock = log10(avgEstConcMock))%>%
+  mutate(logVirus = log10(avgEstConcVirus))%>%
+  mutate(MockSubtracted = logVirus - logMock)
 
@@
 
-Since the 1:2 diluted samples were more likely to be in the detection range, I am only showing data for those.
+## Caveats
 
-## Caveats
-Samples omitted from the plots include:
-
-* Concentrations below the fit curve range.
-Values cannot be extrapolated when the data was *below* the fit curve range.
-However,the MSD software can extrapolate values that are *above* the fit curve range if the curve is linear at the top.
-
-* Samples for which there was only data for a Mock sample and neither of the virus conditions.
-
-Also note:
 If a sample had one rep that was below detection and another that was in range, I used the remaining "in range" rep as the "average" for that sample.

### 2016-09-21 MSD_panels_1_2_3_explant_sup_analysis.Rmd
Changed the way I was subtracting the mock so I could plot on a linear scale, added a plot of just the microarray-interesting subset of donors.

@@
 title: "MSD panels 1, 2 and 3 Explant Supernatant"
-output: pdf_document
+output: html_document
 ---
@@
 
-cat("\n\n\\pagebreak\n")
+#("\n\n\\pagebreak\n")
 
@@
 
-#Do the subtraction
+#When the experimental and mock samples were both "below", we get zero in the minusMockEstConc column, and that won't show up on a log scale.
+So I'm setting all zeros in the minusMockEstConc column to 0.01.
 
-mergeMock<- mergeMock %>%
-  mutate(minusMockEstConc = avgEstConcVirus-avgEstConcMock)
+#Also, in cases where the Mock values are > Virus values, we would get a negative number for the result and not be able to plot it on a log scale.
+So I will take the log of each, then subtract, then plot on a linear scale, so I can have the negatives there.
 
-#When the experimental and mock samples were both "below", we get zero in the minusMockEstConc column, and that won't show up on a log scale.
-So I'm setting all zeros in the minusMockEstConc column to 0.01
+#Recall that log(a/b) = log(a)-log(b)
+#so if Mock = 100 and V = 10, log 100/10 = log (0.1) = -1
+# and log(100)-log(10) = 1-2 = -1
 
 mergeMock <- mergeMock %>%
-  mutate(minusMockEstConc = ifelse(minusMockEstConc <= 0, 0.01, minusMockEstConc))
+  mutate(logMock = log10(avgEstConcMock))%>%
+  mutate(logVirus = log10(avgEstConcVirus))%>%
+  mutate(MockSubtracted = logVirus - logMock)
 
-anyNA(mergeMock)
+```
+
+```{r make adjustments for nice plotting,echo=FALSE}
+
+mergeMock<-mergeMock %>%
+  mutate(Timepoint = ifelse(Timepoint == "T1", "3",
+                            ifelse(Timepoint=="T2","8","24")))
+
+mergeMock<-mergeMock%>%
+  mutate(Virus = ifelse(Virus == "V1", "SD90",
+                        ifelse(Virus == "V2", "186","Mock")))
+
+mergeMock$Timepoint<-factor(mergeMock$Timepoint,levels = c("3","8","24"))
+
 ```
@@
 ## Plots of Analyte Concentrations by Donor, Timepoint and Condition
-We did a 1:5 dilution of the samples so I multiplied the concentrations by 5 to get an estimate of the stock suparnatent concentration.
+We did a 1:5 dilution of the samples so I multiplied the concentrations by 5 to get an estimate of the stock supernatent concentration.
 
-## Caveats
-Samples omitted from the plots include:
-
-* Samples for which there was only data for a Mock sample and neither of the virus conditions (Donor 324, IL-1RA, MIG, RANTES)
-
-Also note:
 If a sample had one rep that was below detection and another that was in range, I used the remaining "in range" rep as the "average" for that sample.
 
-```{r  individual analyte plots, fig.width = 8, fig.height= 13, echo=FALSE}
+```{r  individual analyte plots, fig.width = 6,fig.height= 13, echo=FALSE}
 
-ggplot(mergeMock, aes(x = Timepoint, y = minusMockEstConc))+
+ggplot(mergeMock, aes(x = Timepoint, y = MockSubtracted))+
    geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
-   scale_y_log10()+
+   labs( y = "Estimated Stock Concentration pg/ml")+
+   facet_grid(Assay~Virus, scales = "free_x")+
+   theme(axis.text.x = element_text(size = 11))+
+  #geom_hline(yintercept = 0)+
+  ggtitle("Analyte Concentrations by Donor, Timepoint and Condition")
+
+```
+
+This is just donors 317,318,323 and 324, which were interesting in the microarray
+
+```{r  subset plots, fig.width = 6,fig.height= 13, echo=FALSE}
+
+#Make plots just showing data for donors 317, 318, 323,324
+
+subset<-mergeMock%>%
+  filter(Donor == "317"|Donor == "318"|Donor =="323"|Donor =="324")
+
+#plot it
+
+ggplot(subset, aes(x = Timepoint, y = MockSubtracted))+
+   geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
    labs( y = "Estimated Stock Concentration pg/ml")+
@@
   ggtitle("Analyte Concentrations by Donor, Timepoint and Condition")
-
-```

### 2016-09-21 MSD_panels_1_2_3_explant_sup_analysis.Rmd
minor plot edits

@@
 title: "MSD panels 1, 2 and 3 Explant Supernatant"
-output: html_document
+output: github_document
 ---
@@
    geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
-   labs( y = "Estimated Stock Concentration pg/ml")+
+   labs( y = "log 10 Estimated Stock Concentration pg/ml")+
    facet_grid(Assay~Virus, scales = "free_x")+
@@
   #geom_hline(yintercept = 0)+
-  ggtitle("Analyte Concentrations by Donor, Timepoint and Condition")
+  ggtitle("Analyte Concentrations")
 
@@
    geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
-   labs( y = "Estimated Stock Concentration pg/ml")+
+   labs( y = "log10 Estimated Stock Concentration pg/ml")+
    facet_grid(Assay~Virus, scales = "free_x")+
    theme(axis.text.x = element_text(size = 11))+
-  ggtitle("Analyte Concentrations by Donor, Timepoint and Condition")
+  ggtitle("Analyte Concentrations")

### 2016-09-21 plate_1_TEST_analysis.Rmd
ignore

---
title: "TEST_analysis"
author: "Claire Levy"
date: "June 28, 2016"
output:
  md_document:
    variant: markdown_github

---

This is an analysis of test data that we got from running HVTN clinic CVL samples on "plate 1" See explant_and_HVE_cell_sup_MSD_README for more on the different plates/analyte panels. This plate has the following analytes measured in each well

Plate 1: U-PLEX Biomarker Group 1 (hu) Assays

	U-PLEX Human IFN-γ
	U-PLEX Human IL-1β
	U-PLEX Human IL-2
	U-PLEX Human IL-6
	U-PLEX Human IL-8
	U-PLEX Human IL-10
	U-PLEX Human IL-12p70
	U-PLEX Human TNF-α

The MSD machine gives the data in an excel or text format. I have .csv file with the results.

The definition of each column is in appendix 12 of the Discovery Workbench software manual. Since there are 2 replicates for each sample, the software caluclates an average for some of the measurements (signal, concentration). In order to show both the mean values AND the values for each replicates, the mean values are given twice, once on the row with data for the first rep, then again on the row for its pair.

When FC analyzed the data before she made the following plots for each analyte:

plot of standard curve with upper and lower detection limits (signal on x axis and calculated conc on y)

plot of signal vs dilution (1 10 and 100) for each PTID, including lines showing the upper and lower limits for that analyte.

This is a plot of standard curves for each assay. They are made by plotting the known concentration of the standards (calibrators) vs the signal readout from the machine. These curves are used to extrapolate signal values, and generate "calculated concentrations" of the standards.

```{r standards, echo=FALSE, message=FALSE, warning=FALSE}

library(dplyr)
library(ggplot2)
library(pander)
library(reshape2)

#read in the .csv raw results file
#../ to go up one level and then back down into a different folder.

dat<- read.csv("../raw_data/Plate 1-TEST.csv")

#you get a lot of columns of data
colnames(dat)

#descriptions of what is in each column are in appendix 12 of the Discovery Bench software manual.

# We entered the PTId for "Sample.Group" and the "Sample" is just an arbitrary value

standards <- dat %>%
  filter(Sample.Group == "Standards")

#the input concentrations vs signal. These curves are used to extrapolate the "calculated concentrations"

ggplot(standards, aes(x = Concentration, y = Signal ))+
  geom_point(aes(color = Assay))+
  geom_line(aes(group = Assay, color = Assay))+
  scale_y_log10(breaks = c(10^3, 10^4, 10^5, 10^6))+
  scale_x_log10(breaks = c(10^0, 10^1, 10^2, 10^3, 10^4, 10^5))
```

Of the `r length(standards)` standards that we measured, `r length(standardsInRange)` are in the detection range.  The following samples were NOT in the detection range:

```{r detection range, echo=FALSE, message=FALSE, warning=FALSE}

standardsInRange <-standards%>%
  filter(Detection.Range == "In Detection Range")

standardsOutOfRange<-standards%>%
  filter(Detection.Range != "In Detection Range")%>%
  arrange(Assay,Detection.Range)%>%
  select(Sample.Group,Sample, Assay, Detection.Range)

#only want to keep the standardsInRange

pander(standardsOutOfRange)
```

Discovery Workbench software manual tells us we can evaluate the quality of the fit by looking at the "%Recovery" column for the standards (=Calculated Conc/known conc x 100). The standard concentrations are known, so if the curve is a good fit, the % Recovery should be between 80-120% of the known (input) concentration values.

There following standards had an average percent recovery outside of the "good fit" (<80 or >120) range

```{r curve fit, echo=FALSE, message=FALSE, warning=FALSE}

#Making a table of the avg % recovery. Sample replicates are given the same entry in the "Sample" column, so they can be grouped that way, and then averaged. Note that some replicates may have been eliminated when I filtered for only samples that are in the detection range. For these "singlicates" the mean of the %recovery will be the same as the % recovery.

standard_avg_percent_recov<- standardsInRange %>%
  group_by(Assay, Sample)%>%
  summarize(avg_percent_recov=mean(X..Recovery))

below80Above120<-standard_avg_percent_recov%>%
  filter(avg_percent_recov <80 | avg_percent_recov >120)

pander(below80Above120)

```

The following samples were outside the detection range, I will exclude them from the rest of the analysis:

```{r experimental, echo= FALSE, warning= FALSE, message = FALSE}

#the experimental data ("expDat") is everything except the standards

expDat <- filter(dat, Sample.Group !="Standards")

#Some of the data are not in the detection range

expDatOutOfRange <- expDat %>%
  filter(Detection.Range != "In Detection Range")%>%
  select(Assay, Sample.Group, Dilution, Detection.Range)

pander(expDatOutOfRange)

#Just keep the ones in the detection range

expDatInRange<-filter(expDat, Detection.Range == "In Detection Range")

#calculate the average for the signal for each Assay + Sample.Group + Dilution condition.

expDatInRangeAvgSignal <- expDatInRange %>%
  group_by(Assay, Sample.Group, Dilution)%>%
  summarize(Avg_Signal = mean(Signal))

ggplot(expDatInRangeAvgSignal, aes(x = Dilution, y = Avg_Signal))+
  geom_point(aes( ))

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2016-09-21 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
update?

@@
 title: "MSD panels 1, 2 and 3 HVE Supernatant"
-output: pdf_document
+output: html_document
 ---
@@
 
-ggplot(avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet, aes(x = Timepoint, y = avgEstConc))+
+ggplot(mergeMock, aes(x = Timepoint, y = MockSubtracted))+
    geom_point(aes(color = Donor), alpha = 0.75, size = 3)+
-  scale_y_log10()+
    labs( y = "Estimated Stock Concentration pg/ml")+
@@
 
-################## another way of making the plots #########
-
-#make a list of the analytes
-
-#I'm using a unique vector here instead of the levels of Assay because I don't want to include IFNa2a OR Il-7.
-There weren't any data points in the detection range for those analytes (so I don't want to try to make a plot of it)
-
-# analyteList<-as.list(unique(avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet$Assay))
-
-# #write a function to make a plot for each analyte
-# 
-# plotAnalytes <- function (Analyte){
-#    avgConc_HVE_all_panels_noStds_1to2_noBelow_fullSet %>%
-#      filter(Assay == Analyte)%>%
-#      
-#      ggplot( aes(x = Timepoint, y = avgEstConc))+
-#    geom_point(aes(color = Donor ), alpha = 0.5, size = 3)+
-#   geom_line(aes(group = Donor, color = Donor), alpha = 0.5)+
-#     stat_summary(fun.y = mean, geom = "point")+
-#     stat_summary(fun.y = mean, geom = "line",aes(group = 1))+
-#    scale_shape_manual(values = c("Above Fit Curve Range" = 1, "Below Detection Range" = 1, "In Detection Range" = 16))+
-#    scale_y_log10()+
-#    labs( y = "Concentration pg/ml")+
-#   facet_wrap(~Virus, scales = "free_x")+
-#    theme(axis.text.x = element_text(size = 11))+
-#      ggtitle(Analyte)
-# }
-# 
-# 
-# 
-# lapply(analyteList, FUN = plotAnalytes)
-
 ```

### 2016-09-21 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
adjusted final plot with jittering

@@
 title: "MSD panels 1, 2 and 3 HVE Supernatant"
-output: html_document
+output: github_document
 ---
@@
 ggplot(mergeMock, aes(x = Timepoint, y = MockSubtracted))+
-   geom_point(aes(color = Donor), alpha = 0.75, size = 3)+
+   geom_jitter(aes(color = Donor), alpha = 0.75, size = 2.5, width = 0.2, height = 0.2)+
    labs( y = "Estimated Stock Concentration pg/ml")+

### 2016-09-21 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
formatting final plot

@@
 
-```{r  individual analyte plots, echo=FALSE,fig.width=8, fig.height = 10, message = FALSE, warning = FALSE}
+```{r  individual analyte plots, echo=FALSE,fig.width=8, fig.height = 15, message = FALSE, warning = FALSE}
 
@@
    geom_jitter(aes(color = Donor), alpha = 0.75, size = 2.5, width = 0.2, height = 0.2)+
-   labs( y = "Estimated Stock Concentration pg/ml")+
+   labs( y = "log 10 Estimated Stock Concentration pg/ml")+
    facet_grid(Assay~Virus, scales = "free_x")+
    theme(axis.text.x = element_text(size = 11))+
-  ggtitle("Analyte Concentrations by Donor, Timepoint and Condition")
+  ggtitle("Analyte Concentrations ")
 

### 2016-09-21 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
adjust final plot

@@
    theme(axis.text.x = element_text(size = 11))+
-  ggtitle("Analyte Concentrations ")
+  ggtitle("Mock Subtracted Analyte Concentrations ")
 

### 2016-09-21 MSD_panels_1_2_3_explant_sup_analysis.Rmd
trying to get everything updated
edit final plot

--- 2016-09-21 MSD_panels_1_2_3_explant_sup_analysis.Rmd
@@
 
-```{r multiplying by dilution factor, echo=FALSE, message = FALSE, warning = FALSE}
+```{r multiplying by 5, echo=FALSE, message = FALSE, warning = FALSE}
 
@@
   #geom_hline(yintercept = 0)+
-  ggtitle("Analyte Concentrations by Donor, Timepoint and Condition")
+  ggtitle("Mock Subtracted Analyte Concentrations")
 
@@
    geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
-   labs( y = "Estimated Stock Concentration pg/ml")+
+   labs( y = "log10 Estimated Stock Concentration pg/ml")+
    facet_grid(Assay~Virus, scales = "free_x")+
    theme(axis.text.x = element_text(size = 11))+
-  ggtitle("Analyte Concentrations by Donor, Timepoint and Condition")
+  ggtitle("Mock Subtracted Analyte Concentrations")

### 2016-09-22 MSD_panels_1_2_3_explant_sup_analysis.Rmd
edit final plot
Making a heat map of logFC 3hr vs 24hrs

@@
 explant_all_panels_noStds_AdjBelow <- explant_all_panels_noStds %>%
-mutate(Estimated_Stock_Conc = ifelse(str_detect(Detection.Range,"Below"),0.01, Estimated_Stock_Conc))
+mutate(Estimated_Stock_Conc = ifelse(str_detect(Detection.Range,"Below"),0.001, Estimated_Stock_Conc))
 
@@
 * Note that this is on a log(10) scale; most concentrations are a lot lower than the high ones.
+
+```{r make adjustments for nice plotting, echo=FALSE, message = FALSE, warning = FALSE}
+
+explant_all_panels_noStds_AdjBelow<-explant_all_panels_noStds_AdjBelow %>%
+  mutate(Timepoint = ifelse(Timepoint == "T1", "3",
+                            ifelse(Timepoint=="T2","8","24")))
+
+#factor to get order right
+explant_all_panels_noStds_AdjBelow$Timepoint<-factor(explant_all_panels_noStds_AdjBelow$Timepoint,levels = c("3","8","24"))
+
+explant_all_panels_noStds_AdjBelow<-explant_all_panels_noStds_AdjBelow%>%
+  mutate(Virus = ifelse(Virus == "V1", "SD90",
+                        ifelse(Virus == "V2", "186","Mock")))
+
+explant_all_panels_noStds_AdjBelow$Virus<-factor(explant_all_panels_noStds_AdjBelow$Virus,levels = c("Mock","SD90","186"))
+
+```
 
@@
 
-```{r  subtract mock, fig.width = 8, fig.height= 13, echo=FALSE}
+```{r  merge and subtract mock, fig.width = 8, fig.height= 13, echo=FALSE}
 
@@
 mock <- avgConc_explant_all_panels_noStds_AdjBelow %>%
-  filter(Virus == "M")
+  filter(Virus == "Mock")
 
@@
   avgConc_explant_all_panels_noStds_AdjBelow %>%
-  filter(Virus !="M")
+  filter(Virus !="Mock")
 
@@
 
-#When the experimental and mock samples were both "below", we get zero in the minusMockEstConc column, and that won't show up on a log scale.
-So I'm setting all zeros in the minusMockEstConc column to 0.01.
-
-#Also, in cases where the Mock values are > Virus values, we would get a negative number for the result and not be able to plot it on a log scale.
-So I will take the log of each, then subtract, then plot on a linear scale, so I can have the negatives there.
-
-#Recall that log(a/b) = log(a)-log(b)
-#so if Mock = 100 and V = 10, log 100/10 = log (0.1) = -1
-# and log(100)-log(10) = 1-2 = -1
-
 mergeMock <- mergeMock %>%
-  mutate(logMock = log10(avgEstConcMock))%>%
-  mutate(logVirus = log10(avgEstConcVirus))%>%
-  mutate(MockSubtracted = logVirus - logMock)
-
-```
-
-```{r make adjustments for nice plotting,echo=FALSE}
-
-mergeMock<-mergeMock %>%
-  mutate(Timepoint = ifelse(Timepoint == "T1", "3",
-                            ifelse(Timepoint=="T2","8","24")))
-
-mergeMock<-mergeMock%>%
-  mutate(Virus = ifelse(Virus == "V1", "SD90",
-                        ifelse(Virus == "V2", "186","Mock")))
-
-mergeMock$Timepoint<-factor(mergeMock$Timepoint,levels = c("3","8","24"))
+  mutate(mockSubtracted = avgEstConcVirus - avgEstConcMock)
 
@@
 
-```{r  individual analyte plots, fig.width = 6,fig.height= 13, echo=FALSE}
+```{r setup for individual analyte HEATMAP, echo=FALSE}
 
-ggplot(mergeMock, aes(x = Timepoint, y = MockSubtracted))+
-   geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
-   labs( y = "log 10 Estimated Stock Concentration pg/ml")+
-   facet_grid(Assay~Virus, scales = "free_x")+
-   theme(axis.text.x = element_text(size = 11))+
-  #geom_hline(yintercept = 0)+
-  ggtitle("Mock Subtracted Analyte Concentrations")
+#I want to make a heatmap that has the following data
+
+#NO 8hr timepoint data
+
+#tiles represent foldchange of Mock subtracted 24hr data over mock-subtracted 3hr. So if 24hr>3hr, you get a ratio <1 and if 24hr <3hr you get a ratio >1
+
+#Want gradient intensity to max out at FC of 10.
+
+#Exclude 8hr data.
+
+no8mergeMock <- mergeMock %>%
+  filter(Timepoint != "8")
+
+# Extract the 3hr data and select a few columns
+
+timepoint3<- no8mergeMock %>%
+  filter(Timepoint == "3")%>%
+  select(Donor, Assay, Sample, Virus, mockSubtracted)
+
+#extract 24 hr data and select a few columns
+timepoint24<- no8mergeMock %>%
+  filter(Timepoint =="24")%>%
+  select(Donor, Assay, Sample, Virus, mockSubtracted)
+
+```
+
+```{r merge 24 and 3, check NAs, echo=FALSE}
+#merge 24hr and 3hr data
+
+merge24And3<- merge(timepoint24,timepoint3, by  = c("Donor","Assay","Virus"), all.x = TRUE, all.y = TRUE)
+
+#There are 100 rows where we have data for 24hr but NOT for 3hr.
+which24no3<-merge24And3[which(is.na(merge24And3$Sample.y)),]
+
+#there are NO rows where we have data for 3hr but not for 24hr
+which3no24<-merge24And3[which(is.na(merge24And3$Sample.x)),]
+
+#remove the NA's for tidyness
+
+merge24And3 <- na.omit(merge24And3)
+
+#Fix up colnames      
+names(merge24And3)[4:7]<- c("Sample.24", "mockSubtracted.24","Sample.3", "mockSubtracted.3")
+```
+
+```{r adjust zeros, check NAs, echo=FALSE}
+
+#Problems with zeros:
+#If data from both timepoints is zero we'll get NaN for the ratio
+
+#If data from time 3 = 0, we'll get Inf or -Inf
+#If data from time 24 = zero, we'll get zero
+
+#I'm going to change all zeros to 0.001 to avoid these problems #I think this is ok since a heatmap isnt really quantitative
+
+adjZerosMerge24And3<- merge24And3 %>%
+  mutate(mockSubtracted.24 = ifelse(mockSubtracted.24 == 0, 0.001,mockSubtracted.24))%>%
+  mutate(mockSubtracted.3 = ifelse(mockSubtracted.3 == 0, 0.001,mockSubtracted.3))
+
+```
+
+```{r get ratio, check NAs, echo=FALSE}
+
+#Do the division to get a foldchange. Since we are looking at ratios, it doesnt give any info if the number is negative, plus we can't show that on a log scale. 
+#So I'm taking the abs value of the ratio.
+#and also making a column for the log10 of the ratio for plotting.
+
+adjZerosMerge24And3 <- adjZerosMerge24And3 %>%
+  mutate(AbsVal_mockSub24_Over_mockSub3 = abs(mockSubtracted.24/mockSubtracted.3))%>%
+  mutate(log10ratio = log10(AbsVal_mockSub24_Over_mockSub3))
+
+```
+
+```{r logFC heatmap, check NAs, echo=FALSE}
+
+ggplot(adjZerosMerge24And3, aes(x = Donor, y = Assay))+
+  geom_tile(aes(fill = log10ratio), color = "white")+
+  scale_fill_gradient(low = "green", high = "red")+
+  facet_wrap(~Virus)+
+  ggtitle("log fold change of concentrations at 24hrs over 3 hrs")
 
@@
 This is just donors 317,318,323 and 324, which were interesting in the microarray
-
-```{r  subset plots, fig.width = 6,fig.height= 13, echo=FALSE}
-
-#Make plots just showing data for donors 317, 318, 323,324
-
-subset<-mergeMock%>%
-  filter(Donor == "317"|Donor == "318"|Donor =="323"|Donor =="324")
-
-#plot it
-
-ggplot(subset, aes(x = Timepoint, y = MockSubtracted))+
-   geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
-   labs( y = "log10 Estimated Stock Concentration pg/ml")+
-   facet_grid(Assay~Virus, scales = "free_x")+
-   theme(axis.text.x = element_text(size = 11))+
-  ggtitle("Mock Subtracted Analyte Concentrations")

### 2016-09-22 MSD_panels_1_2_3_explant_sup_analysis.Rmd
changed to doing subtracting log virus-log mock and then getting ratio of 24hr  /3hr  by subtraction because loga - logb = log(a/b)

@@
 
+#do the subtraction, with the result being the difference between the log10 of virus - log 10 Mock
 mergeMock <- mergeMock %>%
-  mutate(mockSubtracted = avgEstConcVirus - avgEstConcMock)
+  mutate(logMockSubtracted = log10(avgEstConcVirus) - log10(avgEstConcMock))
 
@@
 
-## Plots of Analyte Concentrations by Donor, Timepoint and Condition
 We did a 1:5 dilution of the samples so I multiplied the concentrations by 5 to get an estimate of the stock supernatent concentration.
@@
 
-#tiles represent foldchange of Mock subtracted 24hr data over mock-subtracted 3hr. So if 24hr>3hr, you get a ratio <1 and if 24hr <3hr you get a ratio >1
+#tiles represent foldchange of Mock subtracted 24hr data over mock-subtracted 3hr. So if 24hr>3hr, you get a ratio <1 and if 24hr <3hr you get a ratio >1.
+
+#recall that log(a)-log(b) = log(a/b) so I'm going to do subtration, not division to get the log ratio FC.
 
@@
   filter(Timepoint == "3")%>%
-  select(Donor, Assay, Sample, Virus, mockSubtracted)
+  select(Donor, Assay, Sample, Virus, logMockSubtracted)
 
@@
   filter(Timepoint =="24")%>%
-  select(Donor, Assay, Sample, Virus, mockSubtracted)
+  select(Donor, Assay, Sample, Virus, logMockSubtracted)
 
@@
 #Fix up colnames      
-names(merge24And3)[4:7]<- c("Sample.24", "mockSubtracted.24","Sample.3", "mockSubtracted.3")
-```
-
-```{r adjust zeros, check NAs, echo=FALSE}
-
-#Problems with zeros:
-#If data from both timepoints is zero we'll get NaN for the ratio
-
-#If data from time 3 = 0, we'll get Inf or -Inf
-#If data from time 24 = zero, we'll get zero
-
-#I'm going to change all zeros to 0.001 to avoid these problems #I think this is ok since a heatmap isnt really quantitative
-
-adjZerosMerge24And3<- merge24And3 %>%
-  mutate(mockSubtracted.24 = ifelse(mockSubtracted.24 == 0, 0.001,mockSubtracted.24))%>%
-  mutate(mockSubtracted.3 = ifelse(mockSubtracted.3 == 0, 0.001,mockSubtracted.3))
-
+names(merge24And3)[4:7]<- c("Sample.24", "logMockSubtracted.24","Sample.3", "logMockSubtracted.3")
 ```
@@
 
-#Do the division to get a foldchange. Since we are looking at ratios, it doesnt give any info if the number is negative, plus we can't show that on a log scale. 
-#So I'm taking the abs value of the ratio.
-#and also making a column for the log10 of the ratio for plotting.
+#Do the subtraction to get a foldchange.
 
-adjZerosMerge24And3 <- adjZerosMerge24And3 %>%
-  mutate(AbsVal_mockSub24_Over_mockSub3 = abs(mockSubtracted.24/mockSubtracted.3))%>%
-  mutate(log10ratio = log10(AbsVal_mockSub24_Over_mockSub3))
-
+merge24And3 <- merge24And3 %>%
+  mutate(logMockSub24_minus_logMockSub3 = logMockSubtracted.24-logMockSubtracted.3)
 ```
@@
 
-ggplot(adjZerosMerge24And3, aes(x = Donor, y = Assay))+
-  geom_tile(aes(fill = log10ratio), color = "white")+
+ggplot(merge24And3, aes(x = Donor, y = Assay))+
+  geom_tile(aes(fill = logMockSub24_minus_logMockSub3), color = "white")+
   scale_fill_gradient(low = "green", high = "red")+

### 2016-09-22 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
get logFC ratio by doing difference of log mock subtracted for 24hr and 3 hr (bcause log(a) -log(b) = log(a/b)

--- 2016-09-21 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
@@
 title: "MSD panels 1, 2 and 3 HVE Supernatant"
-output: github_document
+output: html_document
 ---
@@
 
-```{r filter for 1:2, echo=FALSE, message = FALSE, warning = FALSE}
-### NOTE ### I'm going to filter for just the 1:2 dilution samples here since they gave more in-detection.
+```{r filter for 1:2 MOSTLY, echo=FALSE, message = FALSE, warning = FALSE}
+### NOTE ### I'm going to filter for just the 1:2 dilution samples here since they gave more in-detection. its ca''ed short because theres only 1:2 data
 
 HVE_all_panels_noStds_AdjBelow_short<-HVE_all_panels_noStds_AdjBelow %>%
-  filter(Sample_Dilution =="1:2", Detection.Range != "Above Fit Curve Range")
+  filter(Sample_Dilution =="1:2")
 
@@
 
-HVE_all_panels_noStds_AdjBelow_short <- rbind(HVE_all_panels_noStds_AdjBelow_short,IL1RA_4_T3,IL1RA_2_T2)
+HVE_all_panels_noStds_AdjBelow_short <- HVE_all_panels_noStds_AdjBelow_short %>%
+  filter(!(Assay == "IL1-RA" & Donor == 2 & 
+             Timepoint == "T2" & Sample_Dilution=="1:2")) %>% 
+  filter(!(Assay == "IL1-RA" & Donor == 4 & 
+             Timepoint == "T3" & Sample_Dilution=="1:2"))
+
+HVE_all_panels_noStds_AdjBelow_short <- rbind(
+  HVE_all_panels_noStds_AdjBelow_short, IL1RA_4_T3, IL1RA_2_T2)
+
+```
+
+```{r make adjustments for nice plotting, echo=FALSE, message = FALSE, warning = FALSE}
+
+HVE_all_panels_noStds_AdjBelow_short<-HVE_all_panels_noStds_AdjBelow_short %>%
+  mutate(Timepoint = ifelse(Timepoint == "T1", "3",
+                            ifelse(Timepoint=="T2","8","24")))
+
+#factor to get order right
+HVE_all_panels_noStds_AdjBelow_short$Timepoint<-factor(HVE_all_panels_noStds_AdjBelow_short$Timepoint,levels = c("3","8","24"))
+
+HVE_all_panels_noStds_AdjBelow_short<-HVE_all_panels_noStds_AdjBelow_short%>%
+  mutate(Virus = ifelse(Virus == "V1", "SD90",
+                        ifelse(Virus == "V2", "186","Mock")))
+
+HVE_all_panels_noStds_AdjBelow_short$Virus<-factor(HVE_all_panels_noStds_AdjBelow_short$Virus,levels = c("Mock","SD90","186"))
 
@@
 
-#Result is a df where the samples with dischordant replicates in terms of Detection.Range.
+#Result is a df where the samples with discordant replicates in terms of Detection.Range.
 The samples with "1" in the Number_of_Reps colimn are the ones where I want to use the data from the rep that was in/above range for the avg, NOT the average between the in-range rep and the below-range rep.
@@
 
-```{r  subtract mock, fig.width = 8, fig.height= 13, echo=FALSE}
+```{r merge and subtract mock, fig.width = 8, fig.height= 13, echo=FALSE}
 
@@
 mock <- avgConc %>%
-  filter(Virus == "M")
+  filter(Virus == "Mock")
 
@@
   avgConc %>%
-  filter(Virus !="M")
+  filter(Virus !="Mock")
 
@@
 whichNAs<-mergeMock[which(is.na(mergeMock$Sample.x)),]
-
 #no NA's!
@@
 
-#When the experimental and mock samples were both "below", we get zero in the minusMockEstConc column, and that won't show up on a log scale.
-So I'm setting all zeros in the minusMockEstConc column to 0.01.
-
-#Also, in cases where the Mock values are > Virus values, we would get a negative number for the result and not be able to plot it on a log scale.
-So I will take the log of each, then subtract, then plot on a linear scale, so I can have the negatives there.
-
-#Recall that log(a/b) = log(a)-log(b)
-#so if Mock = 100 and V = 10, log 100/10 = log (0.1) = -1
-# and log(100)-log(10) = 1-2 = -1
-
+#do the subtraction, with the result being the difference between the log10 of virus - log 10 Mock
 mergeMock <- mergeMock %>%
-  mutate(logMock = log10(avgEstConcMock))%>%
-  mutate(logVirus = log10(avgEstConcVirus))%>%
-  mutate(MockSubtracted = logVirus - logMock)
+  mutate(logMockSubtracted = log10(avgEstConcVirus) - log10(avgEstConcMock))
 
@@
 
-## Plots of Analyte Concentrations by Donor, Timepoint and Condition
-
-## Caveats
+Samples are multiplied by their dilution factor to get estimated concentrations.
 
@@
 
-```{r  individual analyte plots, echo=FALSE,fig.width=8, fig.height = 15, message = FALSE, warning = FALSE}
+```{r setup for individual analyte HEATMAP, echo=FALSE}
 
-ggplot(mergeMock, aes(x = Timepoint, y = MockSubtracted))+
-   geom_jitter(aes(color = Donor), alpha = 0.75, size = 2.5, width = 0.2, height = 0.2)+
-   labs( y = "log 10 Estimated Stock Concentration pg/ml")+
-   facet_grid(Assay~Virus, scales = "free_x")+
-   theme(axis.text.x = element_text(size = 11))+
-  ggtitle("Analyte Concentrations ")
+#I want to make a heatmap that has the following data
+
+#NO 8hr timepoint data
+
+#tiles represent foldchange of Mock subtracted 24hr data over mock-subtracted 3hr. So if 24hr>3hr, you get a ratio <1 and if 24hr <3hr you get a ratio >1
+
+#Want gradient intensity to max out at FC of 10.
+
+#Exclude 8hr data.
+
+no8mergeMock <- mergeMock %>%
+  filter(Timepoint != "8")
+
+# Extract the 3hr data and select a few columns
+
+timepoint3<- no8mergeMock %>%
+  filter(Timepoint == "3")%>%
+  select(Donor, Assay, Sample, Virus, logMockSubtracted)
+
+#extract 24 hr data and select a few columns
+timepoint24<- no8mergeMock %>%
+  filter(Timepoint =="24")%>%
+  select(Donor, Assay, Sample, Virus, logMockSubtracted)
 
 ```
+
+```{r merge 24 and 3, check NAs, echo=FALSE}
+#merge 24hr and 3hr data
+
+merge24And3<- merge(timepoint24,timepoint3, by  = c("Donor","Assay","Virus"), all.x = TRUE, all.y = TRUE)
+
+#There are NO rows where we have data for 24hr but NOT for 3hr.
+which24no3<-merge24And3[which(is.na(merge24And3$Sample.y)),]
+
+#there are NO rows where we have data for 3hr but not for 24hr
+which3no24<-merge24And3[which(is.na(merge24And3$Sample.x)),]
+
+#Fix up colnames      
+names(merge24And3)[4:7]<- c("Sample.24", "logMockSubtracted.24","Sample.3", "logMockSubtracted.3")
+```
+
+```{r get ratio, check NAs, echo=FALSE}
+
+#Do the subtraction to get a foldchange.
+
+merge24And3 <- merge24And3 %>%
+  mutate(logMockSub24_minus_logMockSub3 = logMockSubtracted.24-logMockSubtracted.3)
+```
+
+```{r setup logFC heatmap, check NAs, echo=FALSE}
+
+#I want the scale to have max intensity a 10-fold change in either direction (log10(10)=1 on my scale and log10(0.1)= -1 on my scale) so I am going to set all values >=1 as 1 and all values <=-1 as -1.
+
+#set all values >=1 as 1
+
+merge24And3 <- merge24And3 %>%
+  mutate(max10fold =
+           ifelse(
+             logMockSub24_minus_logMockSub3 >=1,
+         1,
+           ifelse(
+             logMockSub24_minus_logMockSub3 <=-1,
+         -1,logMockSub24_minus_logMockSub3)
+           )
+  )
+
+```
+
+```{r logFC heatmap, check NAs, echo=FALSE}
+
+ggplot(merge24And3, aes(x = Donor, y = Assay))+
+  geom_tile(aes(fill = max10fold), color = "gray")+
+  scale_fill_gradient2(low = "green", mid = "white", high = "red", midpoint = 0)+
+  facet_wrap(~Virus)+
+  ggtitle("log fold change of concentrations at 24hrs over 3 hrs")
+```
+
+```{r trying pheatmap, check NAs, echo=FALSE}
+
+#trying with pheatmap
+# library(pheatmap)
+# library(tidyr)
+# 
+# #trying with pheatmap since ggplot is funky looking and I want zero in the middle
+# 
+# #pheatmap need it as numeric matrix and can't facet so I'll do v186 separately from sd90
+# 
+# SD90heatmap<-merge24And3 %>%
+#   filter(Virus == "SD90")%>%
+#   select(Donor, Assay, max10fold)
+# 
+# 
+# matrixSD90heatmapSpread<-SD90heatmap%>%
+#   spread(key = Donor, value = max10fold)%>%
+#   select(-Assay)%>%
+#   as.matrix()
+# 
+# 
+# rownames(matrixSD90heatmapSpread) <-SD90heatmapSpread$Assay
+# 
+# 
+# pheatmap(matrixSD90heatmapSpread)
+#
+
+test = matrix(rnorm(200), 20, 10)
+test[1:10, seq(1, 10, 2)] = test[1:10, seq(1, 10, 2)] + 3
+test[11:20, seq(2, 10, 2)] = test[11:20, seq(2, 10, 2)] + 2
+test[15:20, seq(2, 10, 2)] = test[15:20, seq(2, 10, 2)] + 4
+colnames(test) = paste("Test", 1:10, sep = "")
+rownames(test) = paste("Gene", 1:20, sep = "")
+
+pheatmap(test)
+
+```

### 2016-09-22 MSD_panels_1_2_3_explant_sup_analysis.Rmd
rechecked data, "missing" in heatmap were actually where we didn't have sup.

@@
 
-#Samples that are "Below Detection Range" means that the signal was "statistically indistinguishable from the background or top of the curve"" and ..."The concentrations calculated are inherently unreliable."
+#Samples that are "Below Detection Range" means that the signal was "statistically indistinguishable from the background or top of the curve" and ..."The concentrations calculated are inherently unreliable."
 
@@
 #Plot of all samples and their concentrations
-#Note that this includes the samples that manually set to 0.01
+#Note that this includes the samples that manually set to 0.001
 
@@
 
-```{r get ratio, check NAs, echo=FALSE}
+```{r get ratio, echo=FALSE}
 
@@
 
-```{r logFC heatmap, check NAs, echo=FALSE}
+```{r setup logFC heatmap, echo=FALSE}
 
-ggplot(merge24And3, aes(x = Donor, y = Assay))+
-  geom_tile(aes(fill = logMockSub24_minus_logMockSub3), color = "white")+
-  scale_fill_gradient(low = "green", high = "red")+
-  facet_wrap(~Virus)+
-  ggtitle("log fold change of concentrations at 24hrs over 3 hrs")
+#I want the scale to have max intensity a 10-fold change in either direction (log10(10)=1 on my scale and log10(0.1)= -1 on my scale) so I am going to set all values >=1 as 1 and all values <=-1 as -1.
+
+#set all values >=1 as 1
+
+merge24And3 <- merge24And3 %>%
+  mutate(max10fold =
+           ifelse(
+             logMockSub24_minus_logMockSub3 >=1,
+         1,
+           ifelse(
+             logMockSub24_minus_logMockSub3 <=-1,
+         -1,logMockSub24_minus_logMockSub3)
+           )
+  )
 
@@
 
+```{r logFC heatmap, echo=FALSE}
+
+#Note that we should NOT have data for the following samples  because we didnt have sup for them
+
+# 323-T2-Anything (i.e. no SD90 data)
+# Didnt have sup for 326 T1 OR T2 (and this shows T3/T1 so we don't have anything)
+
+ggplot(merge24And3, aes(x = Donor, y = Assay))+
+  geom_tile(aes(fill = max10fold), color = "gray")+
+  scale_fill_gradient2(low = "green", mid = "white", high = "red", midpoint = 0, name = "log10 Fold Change")+
+  facet_wrap(~Virus, scales = "free_x")+
+  ggtitle("log fold change of concentrations at 24hrs over 3 hrs")
+```
 This is just donors 317,318,323 and 324, which were interesting in the microarray

### 2016-09-22 MSD_panels_1_2_3_explant_sup_analysis.Rmd
knit for github

@@
 ```
-This is just donors 317,318,323 and 324, which were interesting in the microarray

### 2016-09-22 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
knit for github

@@
   geom_tile(aes(fill = max10fold), color = "gray")+
-  scale_fill_gradient2(low = "green", mid = "white", high = "red", midpoint = 0)+
+  scale_fill_gradient2(low = "green", mid = "white", high = "red", midpoint = 0, name = "log10 Fold Change")+
   facet_wrap(~Virus)+
@@
 
-test = matrix(rnorm(200), 20, 10)
-test[1:10, seq(1, 10, 2)] = test[1:10, seq(1, 10, 2)] + 3
-test[11:20, seq(2, 10, 2)] = test[11:20, seq(2, 10, 2)] + 2
-test[15:20, seq(2, 10, 2)] = test[15:20, seq(2, 10, 2)] + 4
-colnames(test) = paste("Test", 1:10, sep = "")
-rownames(test) = paste("Gene", 1:20, sep = "")
-
-pheatmap(test)
-
 ```

### 2016-09-23 MSD_panels_1_2_3_explant_sup_analysis.Rmd
made a new logfc file with different analyte names and order

@@
 ```
+
+Here is the same heat map with some changes to to Assay names and order:
+
+```{r rename analytes logFC heatmap, echo=FALSE}
+
+#make a new column to hold the new names
+merge24And3<-merge24And3 %>%
+  mutate(altAssayNames = ifelse(Assay == "MIG",
+                                "CXCL9",
+                                ifelse(Assay == "IP-10",
+                                       "CXCL10",
+                                       ifelse(Assay == "IL1-RA","IL-1RA",as.character(Assay)))))
+
+#factor the new assay names in the order FH wanted:
+#Note that the FIRST level will be at the bottom of the heat map axis!!!
+
+merge24And3$altAssayNames<-factor(merge24And3$altAssayNames)
+
+merge24And3$altAssayNames<-factor(merge24And3$altAssayNames, levels(merge24And3$altAssayNames)[c(3,14,7,18,16,17,8,6,5,12,19,4,20,15,13,11,9,10,1,2)])
+
+ggplot(merge24And3, aes(x = Donor, y = altAssayNames))+
+  geom_tile(aes(fill = max10fold), color = "gray")+
+  scale_fill_gradient2(low = "green", mid = "white", high = "red", midpoint = 0, name = "log10 Fold Change")+
+  facet_wrap(~Virus, scales = "free_x")+
+  labs( y = "Assay")+
+  ggtitle("log fold change of concentrations at 24hrs over 3 hrs")

### 2016-09-23 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
actually knitting for github
added another logfc heatmap with different assay names and order

@@
 
+Here is the same heat map with some changes to to Assay names and order:
+
+```{r rename analytes logFC heatmap, echo=FALSE}
+
+#make a new column to hold the new names
+merge24And3<-merge24And3 %>%
+  mutate(altAssayNames = ifelse(Assay == "MIG",
+                                "CXCL9",
+                                ifelse(Assay == "IP-10",
+                                       "CXCL10",
+                                       ifelse(Assay == "IL1-RA","IL-1RA",as.character(Assay)))))
+
+#factor the new assay names in the order FH wanted:
+#Note that the FIRST level will be at the bottom of the heat map axis!!!
+
+merge24And3$altAssayNames<-factor(merge24And3$altAssayNames)
+
+merge24And3$altAssayNames<-factor(merge24And3$altAssayNames, levels(merge24And3$altAssayNames)[c(6,5,12,11,3,14,7,18,17,19,15,13,1,16,8,4,20,9,10,2)])
+
+ggplot(merge24And3, aes(x = Donor, y = altAssayNames))+
+  geom_tile(aes(fill = max10fold), color = "gray")+
+  scale_fill_gradient2(low = "green", mid = "white", high = "red", midpoint = 0, name = "log10 Fold Change")+
+  facet_wrap(~Virus, scales = "free_x")+
+  labs( y = "Assay")+
+  ggtitle("log fold change of concentrations at 24hrs over 3 hrs")
+```
+
 ```{r trying pheatmap, check NAs, echo=FALSE}

### 2016-09-23 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
added back in the scatter plots of concentrations

@@
 
+```{r scatter plots of mock subtracted, echo=FALSE,fig.width=8, fig.height = 15, message = FALSE, warning = FALSE}
+
+ggplot(mergeMock, aes(x = Timepoint, y = logMockSubtracted))+
+   geom_jitter(aes(color = Donor), alpha = 0.75, size = 2.5, width = 0.2, height = 0.2)+
+  labs( y = "log 10 Estimated Stock Concentration pg/ml")+
+   facet_grid(Assay~Virus, scales = "free_x")+
+   theme(axis.text.x = element_text(size = 11))+
+  ggtitle("Analyte Concentrations ")
+```
+
 ```{r setup for individual analyte HEATMAP, echo=FALSE}

### 2016-09-23 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
added scatter plots back to HVE file and edited plot titles in both files

--- 2016-09-23 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
@@
 
+```{r scatter plots of mock subtracted, echo=FALSE,fig.width=8, fig.height = 15, message = FALSE, warning = FALSE}
+
+ggplot(mergeMock, aes(x = Timepoint, y = logMockSubtracted))+
+   geom_jitter(aes(color = Donor), alpha = 0.75, size = 2.5, width = 0.2, height = 0.2)+
+  labs( y = "log 10 Estimated Stock Concentration pg/ml")+
+   facet_grid(Assay~Virus, scales = "free_x")+
+   theme(axis.text.x = element_text(size = 11))+
+  ggtitle("Mock Subtracted Analyte Concentrations ")
+```
+
 ```{r setup for individual analyte HEATMAP, echo=FALSE}

### 2016-09-23 MSD_panels_1_2_3_explant_sup_analysis.Rmd
added scatter plots back to HVE file and edited plot titles in both files

@@
 
+```{r scatter plots of mocksubtracted, fig.width = 6,fig.height= 13, echo=FALSE}
+
+ggplot(mergeMock, aes(x = Timepoint, y = logMockSubtracted))+
+   geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
+   labs( y = "log 10 Estimated Stock Concentration pg/ml")+
+   facet_grid(Assay~Virus, scales = "free_x")+
+   theme(axis.text.x = element_text(size = 11))+
+  #geom_hline(yintercept = 0)+
+  ggtitle("Mock Subtracted Analyte Concentrations")
+
+```
+
+This is just donors 317,318,323 and 324, which were interesting in the microarray
+
+```{r subset scatter plots, fig.width = 6,fig.height= 13, echo=FALSE}
+
+#Make plots just showing data for donors 317, 318, 323,324
+
+subset<-mergeMock%>%
+  filter(Donor == "317"|Donor == "318"|Donor =="323"|Donor =="324")
+
+#plot it
+
+ggplot(subset, aes(x = Timepoint, y = logMockSubtracted))+
+   geom_point(aes(color = Donor),position = position_swarm(0.1), alpha = 0.75, size = 3)+
+   labs( y = "log10 Estimated Stock Concentration pg/ml")+
+   facet_grid(Assay~Virus, scales = "free_x")+
+   theme(axis.text.x = element_text(size = 11))+
+  ggtitle("Mock Subtracted Analyte Concentrations")
+```
+
 ```{r setup for individual analyte HEATMAP, echo=FALSE}

### 2016-09-26 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
used Lucas's code to get greek letters to show up in the heatmaps!

@@
   facet_wrap(~Virus, scales = "free_x")+
+  scale_y_discrete(breaks = merge24And3$altAssayNames,labels =merge24And3$altAssayNames)+
   labs( y = "Assay")+

### 2016-09-26 MSD_panels_1_2_3_explant_sup_analysis.Rmd
used Lucas's code to get greek letters to show up in the heatmaps!

@@
    theme(axis.text.x = element_text(size = 11))+
-  #geom_hline(yintercept = 0)+
   ggtitle("Mock Subtracted Analyte Concentrations")
@@
   facet_wrap(~Virus, scales = "free_x")+
+  scale_y_discrete(breaks = merge24And3$Assay,labels =merge24And3$Assay)+
   ggtitle("log fold change of concentrations at 24hrs over 3 hrs")
@@
   labs( y = "Assay")+
+  scale_y_discrete(breaks = merge24And3$altAssayNames,labels =merge24And3$altAssayNames)+
   ggtitle("log fold change of concentrations at 24hrs over 3 hrs")

### 2016-10-03 MSD_panels_1_2_3_explant_sup_analysis.Rmd
added a new heatmap just showing a few analytes

@@
   ggtitle("log fold change of concentrations at 24hrs over 3 hrs")
+
+```
+
+This is a heat map showing the same data as above, but fewer analytes
+```{r analyte subset heatmap, echo=FALSE}
+
+#want a heat map that just shows CXCL9, IL-1a, IL-10 and MIP-1b, in that order.
+
+#I'm making a vector of the ones I want so I can avoid typing the stupid greek letters.
+
+assaysIwant<- levels(merge24And3$altAssayNames)
+assaysIwant<- assaysIwant[c(6,9,17,20)]
+
+analyteSubset<- merge24And3 %>%
+  filter(altAssayNames %in% assaysIwant)
+
+#Recall that the order of the factors well get plotted bottom to top, to the one you want at the BOTTOM of hte axis needs to be in factor positiojn [1]
+analyteSubset$altAssayNames <- factor(analyteSubset$altAssayNames, levels (analyteSubset$altAssayNames) [c(6,9,17,20)])
+
+ggplot(analyteSubset, aes(x = Donor, y = altAssayNames))+
+  geom_tile(aes(fill = max10fold), color = "gray")+
+  scale_fill_gradient2(low = "green", mid = "white", high = "red", midpoint = 0, name = "log10 Fold Change")+
+  facet_wrap(~Virus, scales = "free_x")+
+  labs( y = "Assay")+
+  scale_y_discrete(breaks = merge24And3$altAssayNames,labels =merge24And3$altAssayNames)+
+  ggtitle("log fold change of concentrations at 24hrs over 3 hrs")

### 2016-10-06 MSD_panels_1_2_3_explant_sup_analysis.Rmd
added a note about the max intensities of the heatmaps
added in data from Lamar for explants 304 and 306

--- 2016-09-26 MSD_panels_1_2_3_explant_sup_analysis.Rmd
@@
 
-```{r  samples in sets, fig.width = 8, fig.height= 13, echo=FALSE}
+```{r  samples in sets, echo=FALSE}
 
@@
 
+Note that in these heat maps, the maximum intensity is a fold change of 1 or -1. Any values that were above 1 were set to 1 and any below -1 were set to -1.
+
 ```{r setup logFC heatmap, echo=FALSE}
@@
   ggtitle("log fold change of concentrations at 24hrs over 3 hrs")
+
+```
+
+```{r to combine w/ LMF data: analyte subset logFC heatmap, echo=FALSE}
+#want a heat map that just shows CXCL9, IL-1a, IL-10 and MIP-1b, in that order.
+
+#I'm making a vector of the ones I want so I can avoid typing the stupid greek letters.
+
+assaysIwant<- levels(merge24And3$altAssayNames)
+assaysIwant<- assaysIwant[c(6,9,17,20)]
+
+analyteSubset<- merge24And3 %>%
+  filter(altAssayNames %in% assaysIwant)
+
+#Recall that the order of the factors well get plotted bottom to top, to the one you want at the BOTTOM of hte axis needs to be in factor positiojn [1]
+analyteSubset$altAssayNames <- factor(analyteSubset$altAssayNames, levels (analyteSubset$altAssayNames) [c(6,9,17,20)])
+```
+
+```{r adding Lamar data set up, echo=FALSE}
+
+#I have data for MIG (CXCL9), IL-1a, IL-10 and MIP-1b from two explant donors that Lamar did MSD on but I didn't: #304 and 306.
+
+#she sent data in a spreadsheet called "HSV data Explant sups"
+#Data for each analyte is in a different tab and it is in a different format in each tab. There are notes on in this here:
+#J:\MacLabUsers\HLADIK SHARED\Projects\Herpes study\MSD\explant_and HVE_cell_sups\LMF explant 304 306 only\LMF explant 304 306 only README
+
+#####Important note ####
+#For these explants only, V1 = 186 and V2 = SD90.
+FH examined the data and we decided that this must be the case.
+These experiments were done pre-microarray so the assignment must have changed at somepoint.
+See README in J:\MacLabUsers\HLADIK SHARED\Projects\Herpes study\MSD\explant_and HVE_cell_sups\LMF explant 304 306 only
+
+#read in CL compiled Lamar data:
+
+only_304_306<- read.csv("../LMF explant 304 306 only/HSV data Explant sups_CL_compiled.csv")
+
+#make a virus column as my data, except remember that in this case V1 = 186, NOT SD90
+
+only_304_306 <- only_304_306 %>%
+  mutate(Virus = 
+           ifelse (str_detect(SampleName, "V1"),"V1",
+ ifelse(str_detect(SampleName, "V2"), "V2",
+                                ifelse(str_detect(SampleName, "M"),"M",
+                                       ifelse(str_detect(SampleName, "186"), "V1",
+
+ifelse(str_detect(SampleName, "SD90"), "V2", "0"))))))
+
+#fix the timepoints
+only_304_306 <- only_304_306 %>%
+  mutate(Timepoint =
+           ifelse(str_detect(SampleName, "T1"),"T1",
+ifelse(str_detect(SampleName, "T2"), "T2",
+                                ifelse(str_detect(SampleName, "T3"), "T3",
+                                       ifelse(str_detect(SampleName, "3HR"), "T1",
+
+ifelse(str_detect(SampleName, "8HR"), "T2", "T3"))))))
+
+#Fix the dose
+
+only_304_306 <- only_304_306 %>%
+  mutate(Dose =
+           ifelse(str_detect(SampleName, "D1"),"D1",
+ifelse(str_detect(SampleName, "D2"), "D2",
+                                ifelse(str_detect(SampleName, "H[Ii]"), "D1", "D2"))))
+
+#filter out the D2 samples since we are only interested in D1 and also filter out the T2 samples (8hr timepoint) since we are only going to be comparing timepoint 1 and 3 (3 and 24hr)
+
+only_304_306 <- only_304_306 %>%
+  filter(Dose == "D1", Timepoint != "T2")
+```
+
+```{r LMF data fixing missing data and new sample name,echo=FALSE}
+#There are several samples with missing data for the calcConc, presumeably because the ECL was below LOD. Those samples are:
+
+#304-T1-V1-D1 MIP1b
+#304-T1-V1, 304-T1-V2, 304-T1-M IL-1a
+#304-T1-M IL10
+
+#I will remove the rows with these samples, AND ALSO the following samples, because they depend on these missing ones for either mock subtraction or T1 mocksubtracted vs T3 mocksubtracted comparison:
+
+#all other samples (i.e. T3) for 304 IL-1a
+#all other samples for 304 IL-10
+
+no_missing_only_304_306 <- only_304_306 %>%
+  na.omit() %>% #remove the missing samples
+  filter(!(Donor == "304" & Assay == "IL-1a"))%>%
+  filter(!(Donor == "304" & Assay == "IL-10"))
+
+#remove some columns we don't need any more
+#add a new "sample" column by pasting donor, timepoint, virus
+no_missing_only_304_306<- no_missing_only_304_306 %>%
+  select(-c(Sample,SampleName, avgECL))%>%
+  mutate(Sample = paste(Donor,"-",Timepoint,"-",Virus, sep = ""))
+
+```
+
+```{r LMF data multiply by dil factor,echo=FALSE}
+#I want to multiply the averaged analyte concentration (CalcConc)by the dilution factor to get the average ESTIMATED concentration(avgEstConc). Then remove the columns I don't need anymore
+
+no_missing_only_304_306 <- no_missing_only_304_306 %>%
+  mutate(avgEstConc = CalcConc * DilFactor) %>%
+  select(-c(CalcConc, DilFactor))
+
+```
+
+```{r LMF data samples in sets, echo=FALSE}
+
+#Looking at the number of samples in each sample "set"(Same Donor and Timepoint).
+
+no_missing_only_304_306_numberInSets<-no_missing_only_304_306 %>%
+         group_by(Donor,Timepoint,Assay)%>%
+  select(Sample,Donor, Timepoint,Assay)%>%
+  summarize(numInSet = n())
+
+#one set has 2 members because I am missing 304-T1-V1 for MIP1b. This won't affect the mock subtraction or T1 T3 comparison so I can leave it as is.
+
+```
+
+```{r LMF data extract and merge mock, echo=FALSE}
+#I want to subtract out the mock data from the corresponding v1 and V2 data.
+
+#extract out Mock samples
+
+mock_304_306 <- no_missing_only_304_306 %>%
+  filter(Virus == "M")
+
+# remove mock from the rest of the data
+
+no_mock_304_306 <-
+  no_missing_only_304_306 %>%
+  filter(Virus !="M")
+
+merge_304_306_mock<-merge(no_mock_304_306, mock_304_306, by = c("Donor", "Assay", "Timepoint"), all.x = TRUE, all.y = TRUE)
+
+#check for na's 
+#anyNA(merge_304_306_mock)
+#ok!!
+```
+
+```{r LMF data fix column names and do mock subtraction, echo=FALSE}
+
+merge_304_306_mock <- merge_304_306_mock %>%
+  select(-c(Dose.x, Dose.y, Sample.y, Virus.y))
+
+names(merge_304_306_mock)[4:7]<-c("Virus","Sample", "avgEstConcVirus", "avgEstConcMock")
+
+#do mock subtraction and logFC
+merge_304_306_mock<-merge_304_306_mock%>%
+  mutate(logMockSubtracted = log10(avgEstConcVirus) - log10(avgEstConcMock))
+
+```
+
+```{r LMF logFC sub, echo=FALSE}
+# Extract the 3hr data and select a few columns
+
+timepoint3_304_306<- merge_304_306_mock %>%
+  filter(Timepoint == "T1")%>%
+  select(Donor, Assay, Sample, Virus, logMockSubtracted)
+
+timepoint24_304_306<- merge_304_306_mock %>%
+  filter(Timepoint == "T3")%>%
+  select(Donor, Assay, Sample, Virus, logMockSubtracted)
+
+#merge 24hr and 3hr data
+
+merge24And3_304_306<- merge(timepoint24_304_306,timepoint3_304_306, by  = c("Donor","Assay","Virus"), all.x = TRUE, all.y = TRUE)
+
+#There is an NA where I have 304-T3-V1 but was dont have data for 304-T1-V1. This is ok, I know about this missing data. I will omit the row here.
+
+merge24And3_304_306<-na.omit(merge24And3_304_306)
+
+#Fix up colnames      
+names(merge24And3_304_306)[4:7]<- c("Sample.24", "logMockSubtracted.24","Sample.3", "logMockSubtracted.3")
+
+#Do the subtraction to get a foldchange.
+
+merge24And3_304_306 <- merge24And3_304_306 %>%
+  mutate(logMockSub24_minus_logMockSub3 = logMockSubtracted.24-logMockSubtracted.3)
+
+#I want the scale to have max intensity a 10-fold change in either direction (log10(10)=1 on my scale and log10(0.1)= -1 on my scale) so I am going to set all values >=1 as 1 and all values <=-1 as -1.
+
+#set all values >=1 as 1
+
+merge24And3_304_306 <- merge24And3_304_306 %>%
+  mutate(max10fold =
+           ifelse(
+             logMockSub24_minus_logMockSub3 >=1,
+         1,
+           ifelse(
+             logMockSub24_minus_logMockSub3 <=-1,
+         -1,logMockSub24_minus_logMockSub3)
+           )
+  )
+```
+
+```{r LMF data fix the virus and assay names, echo=FALSE}
+
+#remember that for these data, V1 = 186
+merge24And3_304_306 <-merge24And3_304_306 %>%
+  mutate(Virus = ifelse(Virus == "V1","186","SD90"))
+
+#fix assay names
+
+merge24And3_304_306 <-merge24And3_304_306 %>%
+  mutate(altAssayNames =
+ifelse(Assay == "MIG", "CXCL9",
+ifelse(Assay == "IL-1a","IL-1\U03B1",
+ifelse(Assay == "MIP1b", "MIP-1\U03B2", as.character(Assay)))))
+```
+
+```{r combine LMF data and my assay subset data, echo=FALSE}
+#Make donor, virus assay and sample into factors
+merge24And3_304_306$Donor<-factor(merge24And3_304_306$Donor)
+
+merge24And3_304_306$Virus<-factor(merge24And3_304_306$Virus)
+
+merge24And3_304_306$altAssayNames<-factor(merge24And3_304_306$altAssayNames)
+merge24And3_304_306$Sample.24<-factor(merge24And3_304_306$Sample.24)
+
+merge24And3_304_306$Sample.3<-factor(merge24And3_304_306$Sample.3)
+
+#rbind my subsetted dataframe and LMF's data
+LMF_CL_explant<-rbind(analyteSubset, merge24And3_304_306)
+
+```
+
+This is a heat map showing the same data as above, but fewer analytes and with Lamar's explant 304 and 306 data added in.
+The gray areas are there because we were missing some data from explant 304 because it was presumably below the LOD (i.e.
+no concentration was given in Lamar's spreadsheet).
+```{r LMF_CL data log FC heatmap, echo=FALSE}
+
+#Note that we are missing some data from explants 304 and 306 because it was presumeably below LOD, I addressed this in the chunk titled "LMF data fixing missing data.."
+
+ggplot(LMF_CL_explant, aes(x = Donor, y = altAssayNames))+
+  geom_tile(aes(fill = max10fold), color = "gray")+
+  scale_fill_gradient2(low = "green", mid = "white", high = "red", midpoint = 0, name = "log10 Fold Change")+
+  facet_wrap(~Virus, scales = "free_x")+
+  labs( y = "Assay")+
+  scale_y_discrete(breaks = merge24And3$altAssayNames,labels =merge24And3$altAssayNames)+
+  ggtitle("log fold change of concentrations at 24hrs over 3 hrs")
+```

### 2016-10-07 MSD_panels_1_2_3_explant_sup_analysis.Rmd
Trying to add in minimal values for the places where we have below LOD values from Lamars data so it will show up on the heat map. In complete and untested!!

@@
 
-#I will remove the rows with these samples, AND ALSO the following samples, because they depend on these missing ones for either mock subtraction or T1 mocksubtracted vs T3 mocksubtracted comparison:
+#FH asked me to set the concentrations for these "below LOD" data to just below the minimum value for those analytes. I'll subtract 10% of the minimum value and set the below LOD to that.
 
-#all other samples (i.e. T3) for 304 IL-1a
-#all other samples for 304 IL-10
+#get the minimum for the different analytes
+MIP1b<-only_304_306 %>%
+  filter(Assay == "MIP1b")
 
-no_missing_only_304_306 <- only_304_306 %>%
-  na.omit() %>% #remove the missing samples
-  filter(!(Donor == "304" & Assay == "IL-1a"))%>%
-  filter(!(Donor == "304" & Assay == "IL-10"))
+lessThanMinMIP1b<-min((na.omit(MIP1b$CalcConc)))-
+  0.1 * (min((na.omit(MIP1b$CalcConc))))
+
+IL1a<-only_304_306 %>%
+  filter(Assay == "IL1a")
+
+lessThanMinIL1a<-min((na.omit(IL1a$CalcConc)))-
+  0.1 * (min((na.omit(IL1a$CalcConc))))
+
+IL10<-only_304_306 %>%
+  filter(Assay == "IL-10")
+
+lessThanMinMIP1b<-min((na.omit(IL10$CalcConc)))-
+  0.1 * (min((na.omit(IL10$CalcConc))))
+
+#just the samples with nas
+
+missing<- only_304_306[rowSums(is.na(only_304_306)) > 0,]
+
+no_missing <- na.omit(only_304_306)
+
+#fix the missing data
+#TEST THIS ####
+
+missing<- missing %>% 
+  mutate(CalcConc = 
+    ifelse(Assay == "MIP1b" &
+            SampleName =="3HR.186.HI",
+            lessThanMinMIP1b,
+         ifelse(Assay == "IL-1a" &
+                  Timepoint == "T1",
+                  lessThanMinIL1a, 
+           ifelse(Assay == "IL-10" &
+                  SampleName == "24HR.Mock.Hi",
+                  lessThanMinIL10,CalcConc))))
+
+#Now combine the missing and no missing data and continue....
 
@@
 
-#one set has 2 members because I am missing 304-T1-V1 for MIP1b. This won't affect the mock subtraction or T1 T3 comparison so I can leave it as is.
+#all sets are complete since I added in values for the missing data above.
 
@@
 ```{r LMF data extract and merge mock, echo=FALSE}
-#I want to subtract out the mock data from the corresponding v1 and V2 data.
+#I want to subtract out the mock data from the corresponding V1 and V2 data.
 

### 2016-10-07 MSD_panels_1_2_3_explant_sup_analysis.Rmd
replaced missing values (below LOD) in LMF's data with miniumum conc for that analyte minus 10% so they would show up on the heat map.

@@
 
-```{r  merge and subtract mock, fig.width = 8, fig.height= 13, echo=FALSE}
+```{r  merge and subtract mock, echo=FALSE}
 
@@
 
+Note we did not have supernatents for:
+
+Donor 323 for SD90, all time points
+
+Donor 326 for SD90 and 186, 3hr and 8hr
+
 ```{r scatter plots of mocksubtracted, fig.width = 6,fig.height= 13, echo=FALSE}
@@
 
-This is just donors 317,318,323 and 324, which were interesting in the microarray
+This is just donors 317,318,323 and 324, which were interesting in the microarray.
 
@@
 
-Note that in these heat maps, the maximum intensity is a fold change of 1 or -1. Any values that were above 1 were set to 1 and any below -1 were set to -1.
-
 ```{r setup logFC heatmap, echo=FALSE}
@@
 
+Note that in these heat maps, the maximum intensity is a fold change of 1 or -1. Any values that were above 1 were set to 1 and any below -1 were set to -1.
+
+Also note we were missing supernatents for:
+
+Donor 323 for SD90, all time points
+
+Donor 326 for SD90 and 186, 3hr and 8hr
+
 ```{r logFC heatmap, echo=FALSE}
@@
 
-# 323-T2-Anything (i.e. no SD90 data)
+# 323-T1-V1, 323-T2-V1, 323-T3-V1 (so no SD90 data for any time point)
 # Didnt have sup for 326 T1 OR T2 (and this shows T3/T1 so we don't have anything)
@@
 IL1a<-only_304_306 %>%
-  filter(Assay == "IL1a")
+  filter(Assay == "IL-1a")
 
@@
 
-lessThanMinMIP1b<-min((na.omit(IL10$CalcConc)))-
+lessThanMinIL10<-min((na.omit(IL10$CalcConc)))-
   0.1 * (min((na.omit(IL10$CalcConc))))
@@
 
+complete_304_306<-rbind(no_missing,missing)
+
 #remove some columns we don't need any more
 #add a new "sample" column by pasting donor, timepoint, virus
-no_missing_only_304_306<- no_missing_only_304_306 %>%
+complete_304_306<- complete_304_306 %>%
   select(-c(Sample,SampleName, avgECL))%>%
@@
 
-no_missing_only_304_306 <- no_missing_only_304_306 %>%
+complete_304_306 <- complete_304_306 %>%
   mutate(avgEstConc = CalcConc * DilFactor) %>%
@@
 
-no_missing_only_304_306_numberInSets<-no_missing_only_304_306 %>%
+complete_304_306_numberInSets<-complete_304_306 %>%
          group_by(Donor,Timepoint,Assay)%>%
@@
 
-mock_304_306 <- no_missing_only_304_306 %>%
+mock_304_306 <- complete_304_306 %>%
   filter(Virus == "M")
@@
 no_mock_304_306 <-
-  no_missing_only_304_306 %>%
+  complete_304_306 %>%
   filter(Virus !="M")
 
-merge_304_306_mock<-merge(no_mock_304_306, mock_304_306, by = c("Donor", "Assay", "Timepoint"), all.x = TRUE, all.y = TRUE)
+merge_304_306_mock<-merge(complete_304_306, mock_304_306, by = c("Donor", "Assay", "Timepoint"), all.x = TRUE, all.y = TRUE)
 
@@
 
-#There is an NA where I have 304-T3-V1 but was dont have data for 304-T1-V1. This is ok, I know about this missing data. I will omit the row here.
-
-merge24And3_304_306<-na.omit(merge24And3_304_306)
-
 #Fix up colnames      
@@
 #I want the scale to have max intensity a 10-fold change in either direction (log10(10)=1 on my scale and log10(0.1)= -1 on my scale) so I am going to set all values >=1 as 1 and all values <=-1 as -1.
+Right now there aren't any values >1 or <-1 but I'm doing this anyway since I did it for my data and want to have the same columns and protocol.
 
@@
 This is a heat map showing the same data as above, but fewer analytes and with Lamar's explant 304 and 306 data added in.
-The gray areas are there because we were missing some data from explant 304 because it was presumably below the LOD (i.e.
-no concentration was given in Lamar's spreadsheet).
+
+Lamar's data did not have average concentrations for the following samples, probably becuse the electronic signal was below the level of detection:
+
+MIP1b: 304-3hr-186
+
+IL-1a: 304-3hr-186, 304-3hr-SD90, 304-3hr-Mock
+
+IL-10: 304-3hr-Mock
+
+In order to reflect on the heat map that these samples had low values, I set the average concentrations to 10% below the minimum  average concentration for the analyte of interest.
+I.e.
+I set the average concentration of MIP1b for sample 304-3hr-186 to 0.076, which is 10% less than the minimum MIP1b average concentration from among all the MIP1b samples where we *did* have data.
+
 ```{r LMF_CL data log FC heatmap, echo=FALSE}

### 2016-10-17 MSD_panels_1_2_3_HVE_sup_analysis.Rmd
Added summary statistics for IFNg concentrations.

@@
   ))
+```
+
+Apparently epithelial cells don't make IFNg, but we did get some signal. Here are the summary statistics for the dilution-corrected IFNg concentrations:
+
+```{r check IFNg concentrations, echo=FALSE, message = FALSE, warning = FALSE}
+##apparently epithelial cells aren't known to make IFNg, so let's check those concentrations
+
+IFNgOnly<-HVE_all_panels_noStds %>%
+  filter(Assay == "IFNg")
+
+summary(IFNgOnly$Estimated_Stock_Conc)
 
